---
title: "The COVID-19 pandemic worsens existing inequalities in the labor force or The COVID-19 pandemic reverses years of progress in making the labor force more equal"
output: 
  html_document:
    theme: united
  word_document:
    reference_docx: "Report_template.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
	warning = FALSE,
  collapse = TRUE,
  fig.align="left")
```

```{r libraries}
### R version
# R version 4.0.4

### R version
# R 4.0 or above
# 
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("girdExtra", repos = "http://cran.us.r-project.org")
if(!require(corrr)) install.packages("corrr", repos = "http://cran.us.r-project.org")
if(!require(Hmisc)) install.packages("Hmisc", repos = "http://cran.us.r-project.org")
if(!require(sjlabelled)) install.packages("sjlabelled", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if(!require(ggtext)) install.packages("ggtext", repos = "http://cran.us.r-project.org")
if(!require(ggrepel)) install.packages("ggrepel", repos = "http://cran.us.r-project.org")
if(!require(doBy)) install.packages("doBy", repos = "http://cran.us.r-project.org")
if(!require(sjmisc)) install.packages("sjmisc", repos = "http://cran.us.r-project.org")
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
if(!require(flextable)) install.packages("flextable", repos = "http://cran.us.r-project.org")
if(!require(scales)) install.packages("scales", repos = "http://cran.us.r-project.org")
if(!require(ggalt)) install.packages("ggalt", repos = "http://cran.us.r-project.org")
if(!require(janitor)) install.packages("janitor", repos = "http://cran.us.r-project.org")

#### packages needed
library(ggalt) # contingency table
library(tidyverse) # several tools for data manipulation
library(haven) # read SPSS, Stata, and SAS data
library(gridExtra) # arrange multiple grid-based plots on a page
library(corrr) # for correlations correlate()
library(Hmisc) # many functions useful for data analysis
library(sjmisc) # useful to set Na's
library(sjlabelled) # get_labels function
library(ggthemes) # themes for ggplot2
library(ggtext) # fancy text
library(ggrepel) # help with the position of the labels on ggplot2 graph
library(doBy) # 'do' something to data stratified 'by' some variables
library(plotly) # interactive plots
library(flextable) # formatting tables 
library(scales) # add scale symbols and colour to axis
library(janitor) # add totals to tabless
library(shiny)
library(widgetframe)

#### Output configurations
options(digits = 3) # decimal points to 3

### adjustments to theme
theme_economist_white <- theme_economist_white(base_size = 12) 

getwd()

```

COVID-19 has struck men's health more severely than women's. [More men died or ended up in hospitals from the virus](https://globalhealth5050.org/the-sex-gender-and-covid-19-project/the-data-tracker/?explore=country&country=England#search). However, COVID-19 has proven to be not solely a health-crisis. The lockdowns and social distancing measures are not affecting everyone equally on socio-economic levels. Responses of businesses and the government to mitigate the impacts of virus, such as flexible working and the job retention scheme, are not effective for all kinds of jobs. Our research (Link to the Research report) shows that the response to the COVID-19 pandemic is worsening the pre-pandemic inequalities in employment between genders in ethnic minorities. We also show that in particular young and working class groups have been hit the hardest by unemployment during the course of the pandemic.

We use the UK's largest study on employment circumstances, the Labour Force Survey (LFS), which draws on data gathered on the working lives of thousands of workers.

### Men's employment rose most steeply

Before the COVID-19 pandemic struck, the official unemployment levels in the UK stood at one of the lowest recorded. After the first lockdown took in effect unemployment levels rose for everyone, but for men most steeply. At its peak in the summer of 2020 about 1 in 19 men and 1 in 21 women could not find a job when they wanted to.

```{r unemployement men/women, fig.width= 8, fig.height=5}
load("../../Data_clean/LFS_clean.rda")

# Unemployment per sex
unem_sex <- LFS_clean %>% 
                filter(AGE >= 18 & AGE <= 64 & ILODEFR <= 2  & QUARTER <= 16) %>%
                group_by(QUARTER, as_label(ILODEFR), SEX) %>%
                summarise(Active = sum(PWT18))  %>% 
            pivot_wider(names_from = `as_label(ILODEFR)`, values_from = Active) %>%
            adorn_totals("col") %>%
            mutate(Rate = `Not employed` / Total) # unemployment rates per gender 

tiplab <-  tibble(SEX = c(rep("Men",16),rep("Women", 16)),
                  QUARTER = c(seq(1:16), seq(1:16)),
                  qlabel = rep(c("Jan-Mar 2019","Apr-Jun 2019","Jul-Sep 2019","Oct-Dec 2019",
            "Jan-Mar 2020", "Feb-Apr 2020", "Mar-May 2020", "Apr-Jun 2020", "May-Jul 2020",
            "Jun-Aug 2020", "Jul-Sep 2020","Aug-Oct 2020", "Sep-Nov 2020", "Oct-Dec 2020",
            "Nov-Jan 2021", "Dec-Feb 2021"),2)) %>% 
            dplyr::arrange(QUARTER)

unem_sex_plot <- unem_sex %>% 
        ggplot(aes(x = QUARTER, y = Rate, col = as_label(SEX))) +
        geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
        geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
        geom_line(size = 1, show.legend = FALSE) +
        # tooltip labels
        geom_point(aes(text = paste("Unemployment:", round(Rate*100,1),"%", 
                                    "\nGender:", as_label(SEX), 
                                    "\nQuarter:", tiplab$qlabel[tiplab$QUARTER == QUARTER])), 
                   shape = 19) + 
        geom_vline(xintercept = 7, linetype="dotted") +

        #scales
        scale_colour_manual(values = c("#DF9216", "#791F83")) +
        scale_y_continuous(breaks = seq(0,0.06, 0.01), limits = c(0, 0.065), labels =
                            scales::percent_format(accuracy = 1L)) +
        scale_x_continuous(breaks=c(1:16),
                         labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                                  "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                                  "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                                  "Nov-\nJan", "Dec-\nFeb")) +
  #  annotation
        labs(x = "", 
             y = "",
             title ="Men’s unemployment rose most steeply",
             subtitle  = "Unemployment rate UK population, 2019/20/21",
             caption = c("Source: UK Labour Force Survey (Person)")) +
        annotate(geom = "text", x = 16, y = 0.053, label = "Men", color= "#DF9216") +
        annotate(geom = "text", x = 16, y = 0.040, label = "Women", color= "#791F83") +
        annotate(geom = "text", x = 2, y = 0.065, label = "2019") +
        annotate(geom = "text", x = 9.5, y = 0.065, label = "2020") +
        annotate(geom = "text", x = 16, y = 0.065, label = "2021") +

        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
              plot.title = 
                element_text(margin = ggplot2::margin(10, 0, 10, 0)),
              axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                         vjust = 0, size = 12),
              axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                         vjust = 2, size = 12),
              legend.position = "none") 

if (knitr::is_html_output()) {
  ggplotly(unem_sex_plot, tooltip = c("text"), width = 800) %>%
  layout(yaxis = list(fixedrange = TRUE),
         xaxis = list(fixedrange = TRUE)) %>%
  config(displayModeBar = FALSE) %>% 
  layout(title = list(text = "Unemployment rate UK population, 2019-2021",
                      xanchor = "left"),
         font  = list(size = 11),
         margin = list(t = 110, pad = 50),
         annotations = list( x = 8.6, y = 0.025, 
                             text = "Start of the COVID-19\nfirst lockdown", 
                             showarrow = F)) 
} else {
  unem_sex_plot +
    # this annotation is placed incorrectly with ggplotly
    annotate(geom = "text", x = 7.1, y = 0.03, label = "Start of the COVID-19\nfirst lockdown", hjust = 0) 
}
```

Economic crises generally affect men's employment worse than women's. For instance because men work in sectors heavily hit by closures. But women are also less likely than men to report themselves unemployed. They more often take up full-time caring roles in the home after job loss, encouraged by the structure of the UK's welfare system.

### A widening ethnic gap in unemployment levels

Men have been impacted more when overall unemployment is considered. But this overall picture conceals that gender impacts vary with age, ethnicity and occupation. The existing gap in unemployment between white and minority-ethnic workers widened after the start of the pandemic, while 2020 started with a gap of 3%, by the end of 2020 this was 5%. The pandemic has thus intensified the pre-pandemic ethnic penalty and especially for women. For example, women from Indian, Pakistani, non-Chinese Asian and other ethnic background suffered a greater increase in unemployment than men from these backgrounds.

```{r unemployment ethnicity, fig.width= 10, fig.height=10}
unem_eth <- LFS_clean %>% 
        filter(AGE >= 18 & AGE <= 64 & ILODEFR <=2  & QUARTER <= 16) %>%
        group_by(QUARTER, as_label(ILODEFR), SEX, ETHUKEUL) %>%
        summarise(Active = sum(PWT18))  %>% 
        pivot_wider(names_from = `as_label(ILODEFR)`, values_from = Active) %>%
        adorn_totals("col") %>%
        mutate(Rate = `Not employed` / Total)

tiplab_eth <- rbind(tiplab,tiplab,tiplab,tiplab,tiplab,tiplab,tiplab,tiplab,tiplab) %>% 
  mutate(ETHUKEUL = c(rep(as_character(as_label(unem_eth$ETHUKEUL)[1]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[2]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[3]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[4]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[5]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[6]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[7]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[8]), 32),
                      rep(as_character(as_label(unem_eth$ETHUKEUL)[9]), 32))) %>% 
  dplyr::arrange(QUARTER, as_factor(ETHUKEUL))

unem_eth_plot <- unem_eth %>% 
        # select data
        select(QUARTER, ETHUKEUL, SEX, Rate) %>% 
        filter(!is.na(Rate & ETHUKEUL)) %>% 
       
         # graphic type and variables 
        ggplot(aes(x = QUARTER, y = Rate, col = as_label(SEX))) +
        geom_vline(xintercept = 4, color = "#c9c9c9") + # grid line between 2019/2020
        geom_vline(xintercept = 15, color = "#c9c9c9") + # grid line between 2020/2021
        geom_line(size = 1, show.legend = FALSE) +
        # tooltip labels
        geom_point(aes(text = paste("Unemployment:", round(Rate*100,1),"%", 
                                    "\nGender:", as_label(SEX), 
                                    "\nQuarter:", tiplab_eth$qlabel[tiplab_eth$QUARTER == QUARTER],
                                    "\nEthnicity:", as_label(ETHUKEUL))), 
                   shape = 19) +
        geom_vline(xintercept = 7, linetype="dotted") +
        facet_wrap(.~as_label(ETHUKEUL), ncol = 3) +
        
        #scales
        scale_colour_manual(values = c("#DF9216", "#791F83")) +
        scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0,0.18), labels =
                             scales::percent_format(accuracy = 1L)) + 
               scale_x_continuous(breaks=c(1:16),
                         labels=c("Jan-\nMar","","Jul-\nSep","",
                                  "Jan-\nMar", "", "Mar-\nMay", "", "May-\nJul",
                                  "", "Jul-\nSep","", "Sep-\nNov", "",
                                  "Nov-\nJan", "")) +
        # annotation
        labs(x = "", 
             y = "",
             title ="The steepest increases in unemployment were among minority ethnic groups",
             subtitle  = "Unemployment rate of UK population, 2019/20/21",
             caption = c("Source: UK Labour Force Survey (Person)")) +
        annotate(geom = "text", x = 2, y = 0.17, label = "2019", size = 3) +
        annotate(geom = "text", x = 9.5, y = 0.17, label = "2020", size = 3) +
        annotate(geom = "text", x = 16, y = 0.17, label = "2021", size = 3) +

        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.title = element_text(face = "bold",
                                        margin = ggplot2::margin(10, 0, 10, 0)),
              plot.subtitle = element_text(size = 12, hjust = 0),
              axis.text = element_text(size = 10),
              legend.position = "top",
              legend.title = element_blank(),
              legend.text = element_text(size=11),
              strip.text = element_text(face = "bold", hjust = 0, size = 11)) +
        guides(color = guide_legend(override.aes = list(size = 4))) 

if (knitr::is_html_output()) {
ggplotly(unem_eth_plot, tooltip=c("text"), width = 1000,
           height = 700) %>%
  layout(yaxis = list(fixedrange = TRUE),
         xaxis = list(fixedrange = TRUE)) %>%
  config(displayModeBar = FALSE) %>% 
  layout(title = list(text = "Unemployment per ethnicity, 2019-2021",
                      xanchor = "left"),
         font  = list(size = 11),
         margin = list(t = 110, r = 100, l = 0),
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y= 1.1))

} else {
  unem_eth_plot 
}
```

### The summer of 2020: very high unemployment among young people

Male under 25-year olds have been hit particularly badly by unemployment. In the summer of 2020 1 in 6 mean and 1 in 8 women could not find a job when traditional workplaces for young people (shops, pubs, restaurants, and entertainment venues) were closed or only open partially.

```{r age, fig.width= 10, fig.height=7}
unem_age <- LFS_clean %>% 
        filter(AGE >= 18 & AGE <= 64 & ILODEFR <=2  & QUARTER <= 16) %>%
        group_by(QUARTER, as_label(ILODEFR), SEX, AGEEUL_2) %>%
        summarise(Active = sum(PWT18))  %>% 
        pivot_wider(names_from = `as_label(ILODEFR)`, values_from = Active) %>%        
        adorn_totals("col") %>%
        mutate(Rate = `Not employed` / Total)

tiplab_age <- rbind(tiplab,tiplab,tiplab,tiplab,tiplab) %>% 
  mutate(AGEEUL_2 = c(rep(as_character(as_label(unem_age$AGEEUL_2)[1]), 32),
                      rep(as_character(as_label(unem_age$AGEEUL_2)[2]), 32),
                      rep(as_character(as_label(unem_age$AGEEUL_2)[3]), 32),
                      rep(as_character(as_label(unem_age$AGEEUL_2)[4]), 32),
                      rep(as_character(as_label(unem_age$AGEEUL_2)[5]), 32))) %>% 
  dplyr::arrange(QUARTER, as_factor(SEX))

unem_age_plot <- unem_age %>%       
        # select data
        select(QUARTER, AGEEUL_2, SEX, Rate) %>%  
        filter(!is.na(Rate))  %>% 
        # arrange(desc(Rate)) %>%

        # graphic type and variables 
        ggplot(aes(x = QUARTER, y = Rate, col= as_label(AGEEUL_2))) +
        geom_vline(xintercept = 4, color = "#c9c9c9") +
        geom_vline(xintercept = 15, color = "#c9c9c9") +
        geom_line(size = 1, show.legend = FALSE) +
        # tooltip labels
        geom_point(aes(text = paste("Unemployment:", round(Rate*100,1),"%", 
                                    "\nGender:", as_label(SEX), 
                                    "\nQuarter:", tiplab_age$qlabel[tiplab_age$QUARTER == QUARTER],
                                    "\nAge:", as_label(AGEEUL_2))), 
                   shape = 19) +
        geom_vline(xintercept = 7, linetype="dotted") +
        facet_wrap(.~as_label(SEX)) +
        
        #scales
        scale_y_continuous(breaks = seq(0,0.20, 0.02), limits = c(0,0.20), labels =
                             scales::percent_format(accuracy = 1L)) +
        scale_x_continuous(breaks=c(1:16),
                         labels=c("Jan-\nMar","","Jul-\nSep","",
                                  "Jan-\nMar", "", "Mar-\nMay", "", "May-\nJul",
                                  "", "Jul-\nSep","", "Sep-\nNov", "",
                                  "Nov-\nJan", "")) +
        scale_color_manual(name = "", values = c("#FC8DB7", "#791F83", "#006295",
                  "#DF9216", "#A3BE41")) +

        # annotation
        labs(x = "", 
             y = "",
             title ="Highest unemployment for young people",
             subtitle  = "Unemployment rate UK population, 2019/20/21",
             caption = c("Source: UK Labour Force Survey (Person)")) +
        annotate(geom = "text", x = 2, y = 0.19, label = "2019") +
        annotate(geom = "text", x = 9.5, y = 0.19, label = "2020") +
        annotate(geom = "text", x = 16, y = 0.19, label = "2021") +
  
        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
              plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
               plot.subtitle = element_text(size = 12, hjust = 0),
              axis.title.x = element_text(margin = ggplot2::margin(t = 5), 
                                          vjust = 0, size = 12),
              axis.title.y = element_text(margin = ggplot2::margin(r = 3), 
                                          vjust = 2, size = 12),
              legend.position = "top",
              legend.key.width = unit(0.5,"cm"),
              legend.margin = ggplot2::margin(0.5, 0, 0.5, 0, "cm"),
              legend.title = element_blank(),
              legend.text = element_text(size=12),
              strip.text = element_text(face = "bold", hjust = 0, size = 12)) +
        guides(size = FALSE, color = guide_legend(override.aes = list(size = 4))) 


  ggplotly(unem_age_plot, tooltip=c("text"), width = 700,
           height = 500) %>%
  layout(yaxis = list(fixedrange = TRUE),
         xaxis = list(fixedrange = TRUE)) %>%
  config(displayModeBar = FALSE) %>% 
  layout(title = list(text = paste('Figure 2: Highest unemployment for young people',
                                    '<br>',
                                    '<sup>',
                                    'Unemployment rate UK population, 2019/20/21',
                                    '</sup>',
                                    '<br>',
                                    '<br>',
                                    '<br>',
                                    '<br>',
                                    '<br>',
                                   '<br>'),
                      xanchor = "left",
                      yanchor = "top",
                      font  = list(size = 18)),
         margin = list(t = 150, l = 0, b = 100, r = 10),
         xaxis = list(showticklabels = TRUE, autotick = TRUE, tickfont = list(size = 11)),
         xaxis2 = list(showticklabels = TRUE, autotick = TRUE, tickfont = list(size = 11)),
         yaxis = list(tickfont = list(size = 11)),
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     yanchor = "top",
                     x = 0.5, y= 1.2,
                     font  = list(size = 14)))

```

### Working class workers faced the most persistent problems into 2021 

Our research also revealed that while some recovery in employment is visible from 2021, those in the working class are facing more persistent problems. These workers in (semi-)routine jobs are essential workers whose crucial roles was highlighted during the pandemic: cleaners, drivers, assemblers, machine operators, porters and messengers, plumbers, shop workers and care workers.

```{r working class employment, fig.width= 10, fig.height=7}
em_class <- LFS_clean %>% 
        filter(AGE >= 18 & AGE <= 64  & QUARTER <= 16) %>%
        group_by(QUARTER, SEX, NSECMJ10_2) %>%
        summarise(pop = sum(PWT18), empl = sum(PWT18[ILODEFR == 1])) %>%
        mutate(Rate = empl / pop) %>%
        filter(NSECMJ10_2 != 6)

tiplab_class <- rbind(tiplab,tiplab,tiplab,tiplab,tiplab) %>% 
  mutate(NSECMJ10_2 = c(rep(as_character(as_label(em_class$NSECMJ10_2)[1]), 32),
                        rep(as_character(as_label(em_class$NSECMJ10_2)[2]), 32),
                        rep(as_character(as_label(em_class$NSECMJ10_2)[3]), 32),
                        rep(as_character(as_label(em_class$NSECMJ10_2)[4]), 32),
                        rep(as_character(as_label(em_class$NSECMJ10_2)[5]), 32))) %>% 
  dplyr::arrange(QUARTER,as_factor(SEX))

em_class_plot <- em_class %>% 
        # select data
        select(QUARTER, NSECMJ10_2, SEX, Rate) %>%  

        # graphic type and variables 
        ggplot(aes(x = QUARTER, y = Rate, col = as_label(NSECMJ10_2))) +
        geom_vline(xintercept = 4, color = "#c9c9c9") +
        geom_vline(xintercept = 15, color = "#c9c9c9") +
        geom_line(size = 1, show.legend = FALSE) +
        # tooltip labels
        geom_point(aes(text = paste("Employment:", round(Rate*100,1),"%", 
                                    "\nGender:", as_label(SEX), 
                                    "\nQuarter:", tiplab_class$qlabel[tiplab_class$QUARTER == QUARTER],
                                    "\nOccupation:", as_label(NSECMJ10_2))), 
                   shape = 19) +
        geom_vline(xintercept = 7, linetype="dotted") +
        facet_wrap(.~as_label(SEX)) +
        
        #scales
        scale_y_continuous(breaks = seq(0.7,1, 0.05), limits = c(0.7,1), labels = scales::percent_format(accuracy = 1L)) + 
        scale_x_continuous(breaks=c(1:16),
                         labels=c("Jan-\nMar","","Jul-\nSep","",
                                  "Jan-\nMar", "", "Mar-\nMay", "", "May-\nJul",
                                  "", "Jul-\nSep","", "Sep-\nNov", "",
                                  "Nov-\nJan", "")) +
        scale_color_manual(name = "", values = c("#FC8DB7", "#791F83", "#006295",
                  "#DF9216", "#A3BE41")) +
        # annotation
        labs(x = "", 
             y = "Percentage of people in employment",
             title ="Employment has not recovered equally for everyone",
             subtitle  = "Employment rate UK population per occupational class, 2019/20/21",
             caption = c("Note: Employment is the proportion of the working age population that is employed
                         Source: UK Labour Force Survey (Person)")) +
        annotate(geom = "text", x = 2, y = 0.98, label = "2019") +
        annotate(geom = "text", x = 9.5, y = 0.98, label = "2020") +
        annotate(geom = "text", x = 16, y = 0.98, label = "2021") +
  
        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.title = element_text(face = "bold",
                                        margin = ggplot2::margin(10, 0, 10, 0)),
              axis.title.y = element_text(margin = ggplot2::margin(r = 3), 
                                           vjust = 2, size = 11), 
              plot.subtitle = element_text(size = 12, hjust = 0),
              legend.position = "top",
              legend.key.width = unit(0.5,"cm"),
              legend.margin = ggplot2::margin(0.5, 0, 0.5, 0, "cm"),
              legend.title = element_blank(),
              legend.text = element_text(size=12),
              strip.text = element_text(face = "bold", hjust = 0, size = 12)) +
        guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4)))

if (knitr::is_html_output()) {
  ggplotly(em_class_plot, tooltip=c("text"), width = 1000,
           height = 700) %>%
  layout(yaxis = list(fixedrange = TRUE),
         xaxis = list(fixedrange = TRUE)) %>%
  config(displayModeBar = FALSE) %>% 
  layout(title = list(text = "Employment rate per occupational class, 2019-2021",
                      xanchor = "left"),
         font  = list(size = 11),
         margin = list(t = 110, l = 0),
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y= 1.1))
} else {
  em_class_plot
}
```

### Who was made redundant?

Redundancies were also particularly high in the working class. Although men in management jobs were at greatest risk of redundancy, the pandemic barely impacted the overall employment in these occupations. The redundancy for women in management roles was very high too. However, women are more likely to be employed in the public sector, in particular health, which have been largely protected against redundancies.

The main reason for the high amount of redundancies seems to be that employers are cutting staff. As non-essential businesses were experiencing long periods of inactivity as a result of COVID-19 closures and other pressures, many employers opted for fewer employees to carry out the available work. The percentage of people who lost their jobs because of workplaces reducing staff shot up rapidly after the first lockdown, peaking at 38% of all redundancies in summer 2020.

```{r redundancy, fig.width= 10, fig.height=7}
red_class <- LFS_clean %>%
        filter(AGE >= 18 & AGE <= 64 & REDUND == 1  & QUARTER <= 16) %>%
        group_by(QUARTER, SEX, NSECMJ10_2) %>%
        summarise(Redund = sum(PWT18)) %>%
        filter(!is.na(NSECMJ10_2) & NSECMJ10_2 != 6)

# this one misses some months, create new tiplab to correctly label
tiplab_red <- red_class %>% 
  mutate(SEX = as_label(SEX)) %>% 
  dplyr::select(QUARTER, SEX, NSECMJ10_2) %>% 
  distinct() %>% 
  left_join(tiplab, by= c("QUARTER", "SEX"))%>% 
  dplyr::arrange(QUARTER, NSECMJ10_2)

red_class_plot <- red_class %>% 
        ggplot(aes(x = QUARTER, y = Redund, col = as_label(NSECMJ10_2))) +
        geom_vline(xintercept = 4, color = "#c9c9c9") +
        geom_vline(xintercept = 15, color = "#c9c9c9") +
        geom_line(size = 1, show.legend = FALSE) +
        geom_point(aes(text = paste("Redundancy:", round(Redund*100,1), 
                                    "\nGender:", as_label(SEX), 
                                    "\nQuarter:", tiplab_red$qlabel[tiplab_red$QUARTER == QUARTER],
                                    "\nOccupation:", as_label(NSECMJ10_2))), 
                   shape = 19) +
        geom_vline(xintercept = 7, linetype="dotted") +
        facet_wrap(~as_label(SEX)) + 
        
        #scales
        scale_x_continuous(breaks=c(1:16),
                         labels=c("Jan-\nMar","","Jul-\nSep","",
                                  "Jan-\nMar", "", "Mar-\nMay", "", "May-\nJul",
                                  "", "Jul-\nSep","", "Sep-\nNov", "",
                                  "Nov-\nJan", "")) +
        scale_color_manual(name = "", values = c("#FC8DB7", "#791F83", "#006295",
                  "#DF9216", "#A3BE41")) +

        
        # annotation
        labs(x = "", 
             y = "Number of employees made redundant",
             title ="Professional and routine occupations experienced more redundancies",
             subtitle  = "Redundancies UK population, 2019/20/21",
             caption = c("Source: UK Labour Force Survey (Person)")) +
        annotate(geom = "text", x = 2, y = 85000, label = "2019") +
        annotate(geom = "text", x = 9.5, y = 85000, label = "2020") +
        annotate(geom = "text", x = 16, y = 85000, label = "2021") +
        
        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.title = element_text(face = "bold",
                                        margin = ggplot2::margin(10, 0, 10, 0)),
              plot.subtitle = element_text(size = 12, hjust = 0),
              axis.title.y = element_text(margin = ggplot2::margin(r = 3), 
                                           vjust = 2, size = 11), 
              legend.position = "top",
              legend.title = element_blank(),
              legend.text = element_text(size = 12),
              strip.text = element_text(face = "bold", hjust = 0, size = 12)) +
        guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) 

if (knitr::is_html_output()) {
  ggplotly(red_class_plot, tooltip=c("text"), width = 1000,
           height = 700) %>%
  layout(yaxis = list(fixedrange = TRUE),
         xaxis = list(fixedrange = TRUE)) %>%
  config(displayModeBar = FALSE) %>% 
  layout(title = list(text = "Redundancies in UK population 2019-2021",
                      xanchor = "left"),
         font  = list(size = 11),
         margin = list(t = 110, l = 0),
         legend = list(orientation = "h",   # show entries horizontally
                     xanchor = "center",  # use center of legend as anchor
                     x = 0.5, y= 1.1))
} else {
  red_class_plot
}
```

It is known that men's employment suffers the most in periods of economic crisis. Our analysis has shown that the COVID-19 pandemic is no different in this respect. Overall, men have been faced with higher unemployment and redundancies. However, the impact of the pandemic varies with age, ethnicity and occupation as well as gender. Rapidly rising rates of job loss because of employers cutting back on staff paint a worrying picture about work during the pandemic. Especially for workers whose experiences were already shaped by intersecting inequalities of gender, class and ethnicity. Yet unemployment levels do not tell the full story. In our next report, we look at precariousness: how precarious has our work become and which groups are most affected?
