---
title: "Track and Trace Inequalities: Exploring the impacts of COVID-19 on women and men’s working lives in the UK"
author: "Luis Torres, Annegreet Veeken and Tracey Warren, University of Nottingham"
date: "30/04/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
	warning = FALSE)

### R version
# R version 4.0.4

### R version
# R 4.0 or above

#### packages needed
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("girdExtra", repos = "http://cran.us.r-project.org")
if(!require(corrr)) install.packages("corrr", repos = "http://cran.us.r-project.org")
if(!require(Hmisc)) install.packages("Hmisc", repos = "http://cran.us.r-project.org")
if(!require(sjlabelled)) install.packages("sjlabelled", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if(!require(ggtext)) install.packages("ggtext", repos = "http://cran.us.r-project.org")
if(!require(ggrepel)) install.packages("ggrepel", repos = "http://cran.us.r-project.org")
if(!require(doBy)) install.packages("doBy", repos = "http://cran.us.r-project.org")
if(!require(sjmisc)) install.packages("sjmisc", repos = "http://cran.us.r-project.org")
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
if(!require(flextable)) install.packages("flextable", repos = "http://cran.us.r-project.org")
if(!require(scales)) install.packages("scales", repos = "http://cran.us.r-project.org")
if(!require(descr)) install.packages("descr", repos = "http://cran.us.r-project.org")
if(!require(janitor)) install.packages("janitor", repos = "http://cran.us.r-project.org")

library(descr) # contingency table
library(tidyverse) # several tools for data manipulation
library(haven) # read SPSS, Stata, and SAS data
library(gridExtra) # arrange multiple grid-based plots on a page
library(corrr) # for correlations correlate()
library(Hmisc) # many functions useful for data analysis
library(sjmisc) # useful to set Na's
library(sjlabelled) # get_labels function
library(ggthemes) # themes for ggplot2
library(ggtext) # fancy text
library(ggrepel) # help with the position of the labels on ggplot2 graph
library(doBy) # 'do' something to data stratified 'by' some variables
library(plotly) # interactive plots
library(flextable) # formatting tables 
library(scales) # add scale symbols and colour to axis
library(janitor) # add totals to tables

#### Output configurations
options(digits = 3) # decimal points to 3
```

## Introduction

The economic upheaval caused by the COVID-19 outbreak has substantial ramifications for gender equality. Previous economic recessions have affected men’s employment the most. However, this is not a typical recession. The impact of lockdowns and social distancing have a large impact on sectors with high female participation. Similarly, the closure of schools, nurseries and the working from home requirement for many workers increased caring responsibilities, which has large impacts on working mothers, but also on fathers taking responsibility for childcare. The early response by governments and flexible working arrangements adopted by businesses may be mitigating these impacts in ways not comparable to previous social and economic crises. 

We explore the impact of Covid-19 on the paid work of women and men in the UK. We ask how the workplaces of women and men, and their types of work, have been impacted as the pandemic effects rolled out. The research asks whether the pandemic is narrowing or reinforcing existing (pre-pandemic) gender inequalities in our ways of working. 


## Methodology

This report draws on the analysis of the 2020/21 releases of the UK Labour Force Survey (LFS). The LFS is the largest study of the employment circumstances of the UK population. Respondents are interviewed for 5 successive waves at 3-monthly intervals and approximately 20% of the sample is replaced every quarter. In 2020, additional non-calendar quarter data have been released as a response to the need of data in the context of the coronavirus pandemic. 

The LFS replace non-responses with data from previous quarters. Responses brought from previous quarter represent a quarter average of 8,470,590. We remove these responses from our analyses to better account for variations between quarters. The resulted economically active population will be considered in the following sections. 

```{r data, echo=FALSE}

load("./Data_clean/LFS_clean.rda")

N_total <- LFS_clean %>%
        filter(AGE >= 18 & AGE <= 64 & IOUTCOME <= 2) %>%
        group_by(as_label(QUARTER)) %>%
        summarise(Sample = n(), Population = sum(PWT18)) 

N_rolled <- LFS_clean %>%
        filter(AGE >= 18 & AGE <= 64 & IOUTCOME == 6) %>%
        group_by(as_label(QUARTER)) %>%
        summarise(Non_Reponse = sum(PWT18))

N_active <- LFS_clean %>% 
        filter(AGE >= 18 & AGE <= 64 & IOUTCOME <= 2 & ILODEFR <=2) %>%
        group_by(as_label(QUARTER)) %>%
        summarise(Econ_active = sum(PWT18))

N_total <- left_join(N_total, N_rolled, by = "as_label(QUARTER)")

N_total <- left_join(N_total, N_active, by = "as_label(QUARTER)")        
        
N_total %>% flextable() %>% theme_vanilla() %>%
        set_caption(caption = "Sample size and population estimates", style = "table") %>%
        bold(bold = TRUE, part = "header") %>%
        set_header_labels(`as_label(QUARTER)` = "Quarter",
                          Population = "Total Population",
                            Non_Reponse = "Non Responses",
                            Econ_active = "Economically Active") %>%
        autofit() %>%
        align_nottext_col(align = "center", header = TRUE, footer = FALSE) %>%
        add_footer_lines("Source: UK Labour Force Survey (Person)") %>%
        color(part = "footer", color = "#666666") %>%
        print(.) # final table for reporting

```


## Findings 

### Gender impact of COVID-19 on employment levels

As previous economic recessions, the impact of COVID-19 on unemployment levels have hit men the most. From August-October 2020 quarter men's employment has shown signs of recovery. However, during the last lockdown these signs are fading for men but recovery remains constant for women.     

```{r unemployment, echo=FALSE}
# Unemployment per sex

unem_sex <- LFS_clean %>% 
                filter(AGE >= 18 & AGE <= 64 & IOUTCOME <= 2 & ILODEFR <=2) %>%
                group_by(QUARTER, SEX, as_label(ILODEFR)) %>%
                summarise(Active = sum(PWT18))  %>% 
            pivot_wider(names_from = `as_label(ILODEFR)`, values_from = Active) %>%
            adorn_totals("col") %>%
            mutate(Rate = `Not employed` / Total) # unemployment rates per gender 

unem_sex %>% 
        ggplot(aes(x = QUARTER, y = Rate, col = as_label(SEX))) +
        geom_line(size = 1, show.legend = FALSE) +
        geom_point(shape = 19) + 
        geom_vline(xintercept = 3, linetype="dotted") +
  
        #scales
        scale_colour_manual(values = c("#DF9216", "#791F83")) +
        scale_y_continuous(breaks = seq(0,0.06, 0.01), limits = c(0,0.06), labels = label_percent()) +
        scale_x_discrete(limits=c("1","2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12"),
                         labels=c("Jan-Mar20", "Feb-Apr20", "Mar-May20", "Apr-Jun20", "May-Jul20", 
                                  "Jun-Aug20", "Jul-Sep20","Aug-Oct20", "Sep-Nov20", "Oct-Dec20",
                                  "Nov20-Jan21", "Dec20-Feb21")) +
        # annotation
        labs(x = "Quarter", 
             y = "Rate",
             title ="Unemployment Rate for Men and Women",
             subtitle  = "UK Propulation between 18-64 years old",
             caption = c("Source: UK Labour Force Survey (Person)")) +
        annotate(geom = "text", x = 12, y = 0.053, label = "Men", color= "#DF9216") +
        annotate(geom = "text", x = 12, y = 0.040, label = "Women", color= "#791F83") +
        
        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.title = element_text(face = "bold",
                                        margin = margin(10, 0, 10, 0),
                                        size = 13),
              axis.text.x=element_text(angle = 90),
              axis.title.x = element_text(margin = margin(t = 5), vjust = 0, size = 10),
              axis.title.y = element_text(margin = margin(r = 3), vjust = 2, size = 10), 
              axis.text = element_text(size = 11),
              legend.position = "none",
              legend.title = element_blank()) 

```

This trend is particularly explained by the job losses in routine and semi routine occupations.


```{r occupation, echo=FALSE}

LFS_clean <- LFS_clean %>% mutate(NSECMJ10_2 = recode_factor(NSECMJ10, 
                                                           `1` = "Higher managerial & professional",
                                                           `2` = "Lower managerial & professional",
                                                           `3` = "Intermediate occupations", 
                                                           `4` = "Small employers & self-employed",
                                                           `5` = "Lower supervisory & technical",
                                                           `6` = "Routine/semi-routine occupations",
                                                           `7` = "Routine/semi-routine occupations", 
                                                           `8` = "Never worked, unemployed, and nec"))  
unem_class <- LFS_clean %>% 
        filter(AGE >= 18 & AGE <= 64 & IOUTCOME <= 2 & ILODEFR <=2) %>%
        group_by(QUARTER, NSECMJ10_2, SEX, as_label(ILODEFR)) %>%
        summarise(Active = sum(PWT18))  %>% 
        pivot_wider(names_from = `as_label(ILODEFR)`, values_from = Active) %>%
        adorn_totals("col") %>%
        mutate(Rate = `Not employed` / Total) 
        

unem_class %>% 
        # select data
        select(QUARTER, NSECMJ10_2, SEX, Rate) %>%  
        filter(!is.na(Rate & as.integer(NSECMJ10_2)) & as.integer(NSECMJ10_2) != 7)  %>% 
        
        # graphic type and variables 
        ggplot(aes(x = QUARTER, y = Rate, col = as_label(SEX))) +
        geom_line(size = 1, show.legend = FALSE) +
        geom_point(shape = 19) +  
        geom_vline(xintercept = 3, linetype="dotted") +
        facet_wrap(.~as_label(NSECMJ10_2), ncol = 3) +
        
        #scales
        scale_colour_manual(values = c("#DF9216", "#791F83")) +
        scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0,0.08), labels = label_percent()) + 
        scale_x_discrete(limits=c("1","2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12"),
                         labels=c("Jan-Mar20", "Feb-Apr20", "Mar-May20", "Apr-Jun20", "May-Jul20", 
                                  "Jun-Aug20", "Jul-Sep20","Aug-Oct20", "Sep-Nov20", "Oct-Dec20",
                                  "Nov20-Jan21", "Dec20-Feb21")) +
        # annotation
        labs(x = "Quarter", 
             y = "Rate",
             title ="Unemployment Rate for Men and Women by Occupation",
             subtitle  = "UK Population between 18-64 years old ",
             caption = c("Source: UK Labour Force Survey (Person)")) +
        
        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.title = element_text(face = "bold",
                                        margin = margin(10, 0, 10, 0),
                                        size = 13),
              axis.text.x=element_blank(),
              axis.title.x = element_text(margin = margin(t = 5), vjust = 0, size = 10),
              axis.title.y = element_text(margin = margin(r = 3), vjust = 2, size = 10), 
              axis.text = element_text(size = 7),
              legend.position = "top",
              legend.title = element_blank(),
              legend.text = element_text(size=10),
              strip.text = element_text(face = "bold", hjust = 0, size = 9)) +
        guides(color = guide_legend(override.aes = list(size = 3))) 

```



```{r income, echo=FALSE}



```


When ethnic background is taken into account, the trend seems less homogeneous. While the overall unemployment trend for people in the white ethic group from the first lockdown in March 2020 remained between 6% and 4%, unemployment rates for people in other groups fluctuated well beyond 6% each quarter. For most groups the COVID-19 pandemic increased the unemployment levels seen before the first lockdown. In this respect, men from mixed and black backgrounds have seen a rise in their unemployment compare to women, while women from other minorities have been hit the hardest. Employment seems to be recovering in the last couple of quarters for most groups, but for mixed/multiple and other minorities.    


```{r ethnicity, echo=FALSE}

# Unemployment per ethnicity and sex

unem_eth <- LFS_clean %>% 
        filter(AGE >= 18 & AGE <= 64 & IOUTCOME <= 2 & ILODEFR <=2) %>%
        group_by(QUARTER, ETHUKEUL, SEX, as_label(ILODEFR)) %>%
        summarise(Active = sum(PWT18))  %>% 
        pivot_wider(names_from = `as_label(ILODEFR)`, values_from = Active) %>%
        adorn_totals("col") %>%
        mutate(Rate = `Not employed` / Total)

unem_eth %>% 
        # select data
        select(QUARTER, ETHUKEUL, SEX, Rate) %>% 
        filter(!is.na(Rate & ETHUKEUL)) %>% 
       
         # graphic type and variables 
        ggplot(aes(x = QUARTER, y = Rate, col = as_label(SEX))) +
        geom_line(size = 1, show.legend = FALSE) +
        geom_point(shape = 19) +  
        geom_vline(xintercept = 3, linetype="dotted") +
        facet_wrap(.~as_label(ETHUKEUL), ncol = 3) +
        
        #scales
        scale_colour_manual(values = c("#DF9216", "#791F83")) +
        scale_y_continuous(breaks = seq(0,0.2, 0.03), limits = c(0,0.2), labels = label_percent()) + 
        scale_x_discrete(limits=c("1","2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12"),
                         labels=c("Jan-Mar20", "Feb-Apr20", "Mar-May20", "Apr-Jun20", "May-Jul20", 
                                  "Jun-Aug20", "Jul-Sep20","Aug-Oct20", "Sep-Nov20", "Oct-Dec20",
                                  "Nov20-Jan21", "Dec20-Feb21")) +
        # annotation
        labs(x = "Quarter", 
             y = "Rate",
             title ="Unemployment Rate for Men and Women by Ethnic Group",
             subtitle  = "UK Population between 18-64 years old ",
             caption = c("Source: UK Labour Force Survey (Person)")) +

        # theme
        theme_economist_white(gray_bg = FALSE) +
        theme(plot.title = element_text(face = "bold",
                                        margin = margin(10, 0, 10, 0),
                                        size = 13),
              axis.text.x=element_blank(),
              axis.title.x = element_text(margin = margin(t = 5), vjust = 0, size = 10),
              axis.title.y = element_text(margin = margin(r = 3), vjust = 2, size = 10), 
              axis.text = element_text(size = 6),
              legend.position = "top",
              legend.title = element_blank(),
              legend.text = element_text(size=10),
              strip.text = element_text(face = "bold", hjust = 0, size = 8)) +
        guides(color = guide_legend(override.aes = list(size = 3))) 

```





```{r age, echo=FALSE}



```















