---
title: "How is COVID-19 impacting women and men’s working lives in the UK?"
subtitle: 'Research Summary 2: How has the risk of precarious work evolved in the COVID-19 UK?'
author: "Luis D. Torres, Tracey Warren and Annegreet Veeken, and the UK Women’s Budget Group"
date: July 2021
output: 
  word_document:
      reference_docx: "Report_template.docx"
      toc: true
      toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
	warning = FALSE,
  collapse = TRUE)
```

```{r libraries}
#### packages needed
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("girdExtra", repos = "http://cran.us.r-project.org")
if(!require(corrr)) install.packages("corrr", repos = "http://cran.us.r-project.org")
if(!require(Hmisc)) install.packages("Hmisc", repos = "http://cran.us.r-project.org")
if(!require(sjlabelled)) install.packages("sjlabelled", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if(!require(ggtext)) install.packages("ggtext", repos = "http://cran.us.r-project.org")
if(!require(ggrepel)) install.packages("ggrepel", repos = "http://cran.us.r-project.org")
if(!require(doBy)) install.packages("doBy", repos = "http://cran.us.r-project.org")
if(!require(sjmisc)) install.packages("sjmisc", repos = "http://cran.us.r-project.org")
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
if(!require(flextable)) install.packages("flextable", repos = "http://cran.us.r-project.org")
if(!require(scales)) install.packages("scales", repos = "http://cran.us.r-project.org")
if(!require(ggalt)) install.packages("ggalt", repos = "http://cran.us.r-project.org")
if(!require(janitor)) install.packages("janitor", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(rpart)) install.packages("caret", repos = "http://cran.us.r-project.org")

library(ggalt) # contingency table
library(tidyverse) # several tools for data manipulation
library(haven) # read SPSS, Stata, and SAS data
library(gridExtra) # arrange multiple grid-based plots on a page
library(corrr) # for correlations correlate()
library(Hmisc) # many functions useful for data analysis
library(sjmisc) # useful to set Na's
library(sjlabelled) # get_labels function
library(ggthemes) # themes for ggplot2
library(ggtext) # fancy text
library(ggrepel) # help with the position of the labels on ggplot2 graph
library(doBy) # 'do' something to data stratified 'by' some variables
library(plotly) # interactive plots
library(flextable) # formatting tables 
library(scales) # add scale symbols and colour to axis
library(janitor) # add totals to tables
library(caret) # machine learning algorithms 
library(rpart) # decision tree 
library(mltools) # one hot encoding 
library(party)
library(tree)
library(randomForest)
```

```{r data, echo=FALSE}

load("../Data_clean/LFS_clean.rda") # this is the base file

## CREATING A SUB-DATASET

d_precariety <- LFS_clean %>% 
  # set variable types
  mutate_at(c("QUARTER", "PWT18", "PIWT18"), ~as.numeric(.)) %>%  
  # Apply basic filters
  filter(AGE >= 18 & AGE <= 64) %>% 
  filter(ILODEFR == 1) %>% # only people in employment
  filter(STAT <= 2) %>% # only employees and self-employed
  filter(STATR <= 2) # only employees and self-employed current job

``` 

# Background

Covid-19 has proven to be not solely a health crisis. Measures adopted by businesses and governments, such as flexible working and the job retention scheme, have shaped the impacts of the pandemic in ways not comparable to previous social and economic crises, while lockdowns and social distancing measures have had an unequal impact on those jobs frequently done by women and men. We set out to examine if Covid-19 is narrowing or reinforcing existing inequalities among women and men in their working lives. Our first research summary focused on unemployment and job loss before and during the pandemic period. This summary examines precarious work.  

Many factors contribute to the rise of precarious work. Typically, economic crisis and its aftermath have been a key driver affecting the risk of precariousness. As employers and employees find themselves operating in a more uncertain context, new hiring take place on the basis of atypical work^[EUROFOUND (2017). Atypical work. Retrieved from: https://www.eurofound.europa.eu/observatories/eurwork/industrial-relations-dictionary/atypical-work]  such as part-time, temporary, fixed-term, casual work, and zero hours contracts among others. Job-seekers tend to accept these contracts, as otherwise they would continue in unemployment.

This research summary aims at understanding how COVID-19 has impacted, if at all, on the precarisation of work in the UK. To answer this question, we use the UK’s largest study on employment circumstances, the Labour Force Survey (LFS), drawing on analysis of the 2019/20/21 releases. We follow the trend pre- and post-pandemic, considering March/May 2020 as the reference point. We trace the pre-pandemic trend up to the first 2019 quarter (Jan/March). We do this to identify if the COVID-19 pandemic is narrowing or reinforcing existing inequalities in the levels of precariousness for workers aged between 18-64.


# Has work become more precarious during the COVID-19 pandemic?

Precarious work involves instability, insecurity, and social and/or economic vulnerability. However, there is no universally accepted definition of precarious work^[Jain, A. & Hassard, J. (2017). Precarious work: definitions, workers affected and OSH consequences. OHS Wiki. Retrieved from:  https://oshwiki.eu/wiki/Precarious_work:_definitions,_workers_affected_and_OSH_consequences]. The European Parliament’s 2017 resolution ^[European Parliament resolution of 4 July 2017 on working conditions and precarious employment (2016/2221(INI))] on working conditions and precarious employment defines it as ‘employment which does not comply with EU, international and national standards and laws and/or does not provide sufficient resources for a decent life or adequate social protection’. In this respect, precarious work is frequently associated with non-standard or atypical employment. This is generally any employment arrangement that is not full-time and permanent. 

We focus our analysis around two dimensions^[Andrea Broughton et al. (2016). Precarious Employment: Patterns, Trends and Policy Strategies in Europe. European Parliament. Retrieved from: https://www.europarl.europa.eu/thinktank/en/document.html?reference=IPOL_STU(2016)587285]:

  * *Employment arrangements*: part-time work, temporary work and atypical employment including flexitime, annualised hours, zero hours contracts, job sharing and on-call working.
  * *Individual characteristics*: gender, ethnicity, and class. 


## Who is in part-time work?

A part-time worker is someone who works fewer hours than a full-time worker. There is no specific number of hours that makes someone full or part-time, but a full-time worker will usually work 35 hours or more a week.

### Slight decline for female workers

Part-time work is prevalent in the UK, especially for women workers who have caring responsibilities. The level of part-time employment among women had been tapering gently downwards before the pandemic hit, while men had seen some increases in part-time working in the decade after the recession of 2008-9. The pandemic period continued the slight decline for women but little change for men. 

```{r part-time sex, fig.width= 8, fig.height=5}

#### by sex 

t_ptwrk <- d_precariety %>% # summary table
  filter(!is.na(FTPTWK)) %>%  # omit no answers and no applly
  group_by(QUARTER, SEX, FTPTWK) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX) %>% 
  # percentage part-time per quarter
  mutate(percentage = count/sum(count)) %>% 
  filter(FTPTWK == 2)

p_ptwrk <- t_ptwrk %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_factor(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.5, 0.05), limits = c(0, 0.55), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 1: Slight decline in part-time work for women",
       subtitle  = "% of part-time UK workers, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 17, y = 0.12, label = "Men", color= "#DF9216", size = 4) +
  annotate(geom = "text", x = 17, y = 0.38, label = "Women", color= "#791F83", size = 4) +
  annotate(geom = "text", x = 7.1, y = 0.25, label = "Start of the COVID-19\nfirst lockdown", hjust = 0, size = 3.5) +
  annotate(geom = "text", x = 2, y = 0.53, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.53, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.53, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "none") 

p_ptwrk # print the plot


```


### Slight decline for white and ethnic-minority groups




```{r part-time ethinicty, fig.width= 8, fig.height=5}

#### by ethnicity and sex

d_precariety <- d_precariety %>% 
  mutate(ETHUKEUL_2 = recode_factor(factor(ETHUKEUL), # recode ethinicity 
                      `1` = "White",
                      `2` = "Minority-ethnic", `3` = "Minority-ethnic", 
                      `4` = "Minority-ethnic", `5` = "Minority-ethnic",
                      `6` = "Minority-ethnic",`7` = "Minority-ethnic",
                      `8` = "Minority-ethnic",`9` = "Minority-ethnic")) %>% as_numeric(., keep.labels = TRUE)

t_ptwrk_seth <- d_precariety %>% # table summary 
  filter(!is.na(FTPTWK)) %>% # omit no answers and no apply
  group_by(QUARTER, ETHUKEUL_2, FTPTWK) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  # percentage part-time per quarter
  mutate(percentage = count/sum(count)) %>% 
  filter(FTPTWK == 2 & !is.na(ETHUKEUL_2))

p_ptwrk_seth <- t_ptwrk_seth %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(ETHUKEUL_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.5, 0.05), limits = c(0, 0.55), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 2: Slight decline in part-time work for all groups",
       subtitle  = "% of part-time UK workers, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 7.1, y = 0.05, label = "Start of the COVID-19\nfirst lockdown", hjust = 0, size = 3) +
  annotate(geom = "text", x = 2, y = 0.53, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.53, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.53, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(override.aes = list(size = 4))) # change the legend box size

p_ptwrk_seth # print the plot

```


### Virtually no change across occupational groups

Part-time jobs are over-concentrated in routine and semi-routine occupations in the UK. Far fewer senior staff work part-time. The pandemic did little to alter this picture, and trends in levels of part-time working remained virtually unchanged across the occupational groupings in 2020. In 2021, the amount of lower supervisory/technical staff working part-time dipped ??


```{r part-time class, fig.width= 9, fig.height=6}

t_ptwrk_sclass <- d_precariety %>% # table summary 
  filter(!is.na(FTPTWK) & NSECMJ10_2 <= 5) %>% # omit no answers and no apply
  group_by(QUARTER, NSECMJ10_2, FTPTWK) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, NSECMJ10_2) %>% 
  # percentage part-time per quarter
  mutate(percentage = count/sum(count)) %>% 
  filter(FTPTWK == 2 & !is.na(NSECMJ10_2))

p_ptwrk_sclass <- t_ptwrk_sclass %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(NSECMJ10_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.5, 0.05), limits = c(0, 0.55), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 3: Part-time trend remains virtually unchanged",
       subtitle  = "% of part-time UK workers, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.53, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.53, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.53, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_ptwrk_sclass # print the plot

```


### Intersectionality matters

The story of relative stability in levels of part-time employment during the pandemic crisis (as seen in Figures 1-3) is led by the experiences of the employed white majority. Separating white from minority-ethnic workers reveals a far more unsettled picture for the latter group. 

```{r part-time mix, fig.width= 10, fig.height=8}

#### sex, ethnicity and class 

t_ptwrk_mix <- d_precariety %>% # table summary 
  filter(!is.na(FTPTWK) & NSECMJ10_2 <= 5) %>% # # omit no answers and no apply
  group_by(QUARTER, SEX, ETHUKEUL_2, NSECMJ10_2, FTPTWK) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX, ETHUKEUL_2, NSECMJ10_2) %>% 
  # percentage part-time per quarter
  mutate(percentage = count/sum(count)) %>% 
  filter(FTPTWK == 2 & !is.na(NSECMJ10_2) & !is.na(ETHUKEUL_2))

p_ptwrk_mix <- t_ptwrk_mix %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(NSECMJ10_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(as_label(ETHUKEUL_2) ~ as_label(SEX)) +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.7, 0.1), limits = c(0, 0.75), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 4: Intersectionality matters for part-time work",
       subtitle  = "% of part-time UK workers, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.74, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.74, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.74, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_ptwrk_mix # print the plot

```


## What are the reasons for part-time work?


### While men cannot find a full-time job, Women do not want it

Part-time jobs are lower paid than full-time jobs, on the whole, and they bring a range of career penalties. Why people work part-time, given the weaker quality of part-time than full-time jobs, is strongly gendered. Other than students, men who work part-time in the UK are largely working part-time involuntarily: they are looking for full-time work but cannot find it. Most female part-timers, when asked why they work part-time, say they do not want a full-time job. This is because they are invariably working part-time to fit around other roles in their lives. 
The pandemic saw a rise in the percentage of male part-timers working part-time involuntarily, but less of a change among the women???    


```{r part-time reason sex, fig.width= 10, fig.height=6}

#### by sex 

t_yptwrk <- d_precariety %>% # table summary 
  filter(FTPTWK == 2) %>% # only those in par-time work
  group_by(QUARTER, SEX, YPTJOB) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(YPTJOB %in% c("3", "4")) # could not find / did not want ft job

p_yptwrk <- t_yptwrk %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(YPTJOB)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,1, 0.1), limits = c(0, 1.05), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 5: While men cannot find a full-time job, Women do not want it",
       subtitle  = "Reasons for part-time work UK population, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 1.04, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 1.04, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 1.04, label = "2021", size = 3.5) +

  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(override.aes = list(size = 4))) # change the legend box size

p_yptwrk # print graph

```


### Ethnic minorities struggle more to find a full-time job


```{r part-time reason ethnicity, fig.width= 10, fig.height=6}

#### by ethnicity 

t_yptwrk_eth <- d_precariety %>% # table summary
  filter(FTPTWK == 2) %>% # only those in par-time work / ETHUKEUL_2 omit no answers and no apply
  group_by(QUARTER, ETHUKEUL_2, YPTJOB) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(YPTJOB %in% c("3", "4") & !is.na(ETHUKEUL_2)) # could not find / did not want ft job

p_yptwrk_eth <- t_yptwrk_eth %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(ETHUKEUL_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(YPTJOB)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,1, 0.1), limits = c(0, 1.05), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 6: Ethnic minorities struggle more to find a full-time job",
       subtitle  = "Reasons for part-time work UK population, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 1.04, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 1.04, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 1.04, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(override.aes = list(size = 4))) # change the legend box size

p_yptwrk_eth


```

### Lower level occupations struggle more to find full-time jobs


```{r part-time reason class, fig.width= 10, fig.height=6}

#### by class

t_yptwrk_class <- d_precariety %>% # table summary 
  filter(FTPTWK == 2 & NSECMJ10_2 <= 5) %>% # only those in par-time work
  group_by(QUARTER, NSECMJ10_2, YPTJOB) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, NSECMJ10_2) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(YPTJOB %in% c("3", "4")) # could not find / did not want ft job

p_yptwrk_class <- t_yptwrk_class %>% # line graph
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(NSECMJ10_2))) +
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(YPTJOB)) + 
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,1, 0.1), limits = c(0, 1.05), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 7: Lower level occupations struggle more to find full-time jobs",
       subtitle  = "Reasons for part-time work UK population, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 1.04, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 1.04, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 1.04, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_yptwrk_class



```

### Minority-ethnic men from most occupations struggle to find a full-time job


```{r part-time reason mix, fig.width= 10, fig.height=8}

#### sex, ethinicity and class 

t_yptwrk_mix <- d_precariety %>% # table summary
  filter(FTPTWK == 2 & NSECMJ10_2 <= 5) %>% # only those in par-time work
  group_by(QUARTER, SEX, ETHUKEUL_2, NSECMJ10_2, YPTJOB) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX, ETHUKEUL_2, NSECMJ10_2) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(YPTJOB %in% c("3") & !is.na(NSECMJ10_2) & !is.na(ETHUKEUL_2)) # could not find

p_yptwrk_mix <- t_yptwrk_mix %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(NSECMJ10_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_grid(as_label(SEX)~ as_label(ETHUKEUL_2)) + 
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,1, 0.1), limits = c(0, 1.05), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 8: Minority-ethnic men from most occupations struggle to find a full-time job",
       subtitle  = "% people who could not find a full-time job, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 1.04, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 1.04, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 1.04, label = "2021", size = 3.5) +

  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 10),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_yptwrk_mix # print the plot




```



## Who is in temporary work?  


### Slight increase in temporary work



```{r temporary sex, fig.width= 8, fig.height=5}

#### by sex 

t_jobtyp <- d_precariety %>% # summary table
  filter(!is.na(JOBTYP)) %>%  # omit no answer and no apply
  group_by(QUARTER, SEX, JOBTYP) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX) %>% 
  # percentage part-time per quarter
  mutate(percentage = count/sum(count)) %>% 
  filter(JOBTYP == 2)

p_jobtyp <- t_jobtyp %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_factor(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.1, 0.01), limits = c(0, 0.11), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 9: Slight increase in temporary work",
       subtitle  = "% of temporary UK workers, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 17, y = 0.043, label = "Men", color= "#DF9216", size = 4) +
  annotate(geom = "text", x = 17, y = 0.064, label = "Women", color= "#791F83", size = 4) +
  annotate(geom = "text", x = 2, y = 0.105, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.105, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.105, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 10),
        legend.position = "none") 

p_jobtyp # print the plot


```


### Fixed contracts are increasing


```{r type temporary sex, fig.width= 10, fig.height=6}

d_precariety <- d_precariety %>% 
  mutate(RESTMR6_2 = recode_factor(factor(RESTMR6), # recode categories in RESTMR6
                                   `1` = "Permanent",
                                   `2` = "Seasonal", `3` = "Seasonal", `4` = "Seasonal", `5` = "Seasonal", `6` = "Seasonal", `7` = "Seasonal", 
                                   `8` = "Fixed contract", `9` = "Fixed contract", `10` = "Fixed contract", `11` = "Fixed contract", `12` = "Fixed contract", `13` = "Fixed contract",
                                   `14` = "Agency", `15` = "Agency", `16` = "Agency", `17` = "Agency", `18` = "Agency", `19` = "Agency",
                                   `20` = "Casual", `21` = "Casual", `22` = "Casual", `23` = "Casual", `24` = "Casual", `25` = "Casual",
                                   `26` = "Other", `27` = "Other", `28` = "Other", `29` = "Other", `30` = "Other", `31` = "Other",
                                   `32` = "Type not stated")) %>% 
  as_numeric(., keep.labels = TRUE) 


#### by sex 

t_restmr2 <- d_precariety %>% # summary table
  filter(JOBTYP == 2) %>%  
  group_by(QUARTER, SEX, RESTMR6_2) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX) %>% 
  # percentage 
  mutate(percentage = count/sum(count)) %>% 
  filter(!is.na(RESTMR6_2) & RESTMR6_2 != 6) # remove "other"

p_restmr2 <- t_restmr2 %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(RESTMR6_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(SEX)) +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.6, 0.1), limits = c(0, 0.65), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 10: Fixed contracts are increasing",
       subtitle  = "Type of temporary job, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.64, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.64, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.64, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_restmr2 # print the plot


```


### Temporary work affects ethnic minorities the most


```{r temporary ethnicity, fig.width= 8, fig.height=5}

#### ethinicity

t_jobtyp_eth <- d_precariety %>% # table summary 
  filter(!is.na(JOBTYP)) %>% # # omit no answer and no apply
  group_by(QUARTER, ETHUKEUL_2, JOBTYP) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  # percentage part-time per quarter
  mutate(percentage = count/sum(count)) %>% 
  filter(JOBTYP == 2 & !is.na(ETHUKEUL_2))

p_jobtyp_eth <- t_jobtyp_eth %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(ETHUKEUL_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.1, 0.01), limits = c(0, 0.105), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 11: Temporary work affects ethnic minorities the most",
       subtitle  = "% of temporary UK workers, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 17, y = 0.044, label = "White", color= "#DF9216", size = 4) +
  annotate(geom = "text", x = 17, y = 0.088, label = "Minority-ethnic", color= "#791F83", size = 4) +
  annotate(geom = "text", x = 2, y = 0.104, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.104, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.104, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 10),
        legend.position = "none",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_jobtyp_eth # print the plot


```

### Small employers account for most temporary workers


```{r temporary class, fig.width= 8, fig.height=6}

#### class

t_jobtyp_class <- d_precariety %>% # table summary 
  filter(!is.na(JOBTYP) & NSECMJ10_2 <= 5) %>% # remove NA's
  group_by(QUARTER, NSECMJ10_2, JOBTYP) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, NSECMJ10_2) %>% 
  # percentage part-time per quarter
  mutate(percentage = count/sum(count)) %>% 
  filter(JOBTYP == 2 & !is.na(NSECMJ10_2))

p_jobtyp_class <- t_jobtyp_class %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(NSECMJ10_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.6, 0.1), limits = c(0, 0.65), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 12: Small employers account for most temporary workers",
       subtitle  = "% of temporary UK workers, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.64, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.64, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.64, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_jobtyp_class # print the plot




```



## What are the reasons for temporary work?


### Pandemic has made harder to find a permanent job


```{r reason temporary sex, fig.width= 10, fig.height=6}

t_whytem <- d_precariety %>% # table summary 
  # only temp
  filter(JOBTYP == 2) %>% # only those in par-time work
  group_by(QUARTER, SEX, WHYTMP6) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(WHYTMP6 %in% c("3", "4")) # could not find / did not want 

p_whytem <- t_whytem %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(WHYTMP6)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.4, 0.05), limits = c(0, 0.45), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 13: Pandemic has made harder to find a permanent job",
       subtitle  = "Reasons for temporary work, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.42, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.42, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.42, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(override.aes = list(size = 4))) # change the legend box size

p_whytem # print graph


```


### Fixed contracts thrive when permanent jobs don't


```{r type could not find permanent job sex, fig.width= 10, fig.height=6}

##### cannot find permanent job by contract type

t_restmr <- d_precariety %>% # table summary 
  filter(JOBTYP == 2) %>% # only those in part-time work
  group_by(QUARTER, RESTMR6_2, RESTMR6, SEX) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(!is.na(RESTMR6) & RESTMR6 %in% c("4", "10", "16", "22")) # could not find

p_restmr <- t_restmr %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(RESTMR6_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(SEX)) +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.24), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 14: Fixed contracts thrive when permanent jobs don't",
       subtitle  = "Could not find a permanent job, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.22, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.22, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.22, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_restmr # print the plot




```

### Fixed contracts and agency work vary the most among groups


```{r type could not find permanent job ethnicity, fig.width= 10, fig.height=6}

##### cannot find permanenet job by contract type

t_restmr_eth <- d_precariety %>% # table summary 
  filter(JOBTYP == 2) %>% # remove NA's
  group_by(QUARTER, RESTMR6_2, RESTMR6, ETHUKEUL_2) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  # percentage
  mutate(percentage = count/sum(count)) %>% 
  filter(!is.na(ETHUKEUL_2) & !is.na(RESTMR6) & RESTMR6 %in% c("4", "10", "16", "22")) # could not find

p_restmr_eth <- t_restmr_eth %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(RESTMR6_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(ETHUKEUL_2)) + 
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.24), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 15: Fixed contracts and agency work vary the most",
       subtitle  = "Could not find a permanent job, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.22, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.22, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.22, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_restmr_eth # print the plot



```

### Fixed contracts are on the rise for all occupational groups


```{r type could not find permanent job class, fig.width= 10, fig.height=7}

##### cannot find permanent job by contract type

t_restmr_class <- d_precariety %>% # table summary 
  filter(JOBTYP == 2 & NSECMJ10_2 <= 5) %>% # remove NA's
  group_by(QUARTER, RESTMR6_2, RESTMR6, NSECMJ10_2) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, NSECMJ10_2) %>% 
  # percentage
  mutate(percentage = count/sum(count)) %>% 
  filter(!is.na(RESTMR6) & RESTMR6 %in% c("4", "10", "16", "22") & NSECMJ10_2 != 3 & !is.na(NSECMJ10_2)) # remove "other" and self-employed

p_restmr_class <- t_restmr_class %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(RESTMR6_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(NSECMJ10_2)) + 
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.3, 0.05), limits = c(0, 0.34), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 16: Fixed contracts are on the rise for all occupational groups",
       subtitle  = "Could not find a permanent job, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.32, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.32, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.32, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", vjust = 0.5, hjust = 0, size = 11)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_restmr_class # print the plot



```


## Who is in Very atypical employment?  


### Flexitime has increased since the pandemic started


```{r very atypical sex, fig.width= 10, fig.height=6}

#### by sex 

t_fled <- d_precariety %>% # table summary 
  filter(QUARTER %in% c(2, 4, 6, 7, 8, 9, 10, 12, 13, 14, 15, 16, 17, 18)) %>% 
  group_by(QUARTER, FLED10, SEX) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, SEX) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(!is.na(FLED10) & FLED10 %in% c(1, 2, 4, 7, 8)) # omit NAs and include only certain categories

p_fled <- t_fled %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(FLED10))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(SEX)) +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.22), labels =
                       scales::percent_format(accuracy = 1)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 17: Flexitime has increased since the pandemic started",
       subtitle  = "Atypical work arrangements, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.21, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.21, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.21, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_fled # print the plot

```

### Atypical employment ethnic gap is widening


```{r very atypical ethnicity, fig.width= 10, fig.height=6}

#### ethnicity

t_fled_eth <- d_precariety %>% # table summary 
  filter(QUARTER %in% c(2, 4, 6, 7, 8, 9, 10, 12, 13, 14, 15, 16, 17, 18)) %>% 
  group_by(QUARTER, FLED10, ETHUKEUL_2) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(!is.na(FLED10) & !is.na(ETHUKEUL_2) & FLED10 %in% c(1, 2, 4, 7, 8)) # remove NA and include only certain categories

p_fled_eth <- t_fled_eth %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(FLED10))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(ETHUKEUL_2)) +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.2), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 18: Atypical employment ethnic gap is widening",
       subtitle  = "Atypical work arrangements, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.005, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.005, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.005, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 11)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_fled_eth # print the plot

```

### Fexible and annualised hours common in high/middle occupations

```{r very atypical class, fig.width= 10, fig.height=7}

#### class

t_fled_class <- d_precariety %>% # table summary 
  filter(QUARTER %in% c(2, 4, 6, 7, 8, 9, 10, 12, 13, 14, 15, 16, 17, 18) & NSECMJ10_2 <= 5) %>% 
  group_by(QUARTER, FLED10, NSECMJ10_2) %>% 
  summarise(count = sum(PWT18)) %>% 
  group_by(QUARTER, NSECMJ10_2) %>% 
  # percentage reason
  mutate(percentage = count/sum(count)) %>% 
  filter(!is.na(FLED10) & FLED10 %in% c(1, 2, 4, 7, 8) & !is.na(NSECMJ10_2) & NSECMJ10_2 != 3) # only certain categories

p_fled_class <- t_fled_class %>% 
  ggplot(aes(x = QUARTER, y = percentage, colour = as_label(FLED10))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(NSECMJ10_2)) +
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.23), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       title ="Figure 19: Fexible and annualised hours common in high/middle occupations",
       subtitle  = "Atypical work arrangements, 2019/20/21",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.22, label = "2019", size = 3.5) +
  annotate(geom = "text", x = 9.5, y = 0.22, label = "2020", size = 3.5) +
  annotate(geom = "text", x = 17, y = 0.22, label = "2021", size = 3.5) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 9),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 10),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 9),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", vjust = 0.5, hjust = 0, size = 11)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size

p_fled_class # print the plot

```


# The Project and its Data

Our project ‘How is COVID-19 impacting women and men’s working lives in the UK?’ is funded by Health Data Research UK, as part of the rapid funding call to use and enrich the data within the Data & Connectivity National Core Study (NCS) capability. This report draws on the analysis of the 2019/20/21 releases of the UK Labour Force Survey (LFS). Respondents are interviewed for five successive waves at three-months intervals and approximately 20% of the sample is replaced every quarter. Four quarters releases are supported in a typical year: Jan-Mar | Apr-Jun | Jul-Sep | Oct-Dec. From 2020, additional non-calendar quarter data have been released in response to the context of the coronavirus pandemic. Besides the four calendar quarters per year, 2020 included another eight releases of data. The last two quarters in 2020 also included data for January and February 2021. Our analyses consider only those in employment (current jobs) between 18-64 years old. 

Table 1 summarises the sample size and population estimates for all the data waves considered in this analysis. 


```{r sample, echo=FALSE}

N_total_2 <- LFS_clean %>%
  filter(AGE >= 18 & AGE <= 64) %>% 
  filter(ILODEFR == 1) %>% # only people in employment
  filter(STAT <= 2) %>% # only employees and self-employed
  filter(STATR <= 2) %>% # only employees and self-employed current job
  group_by(as_label(QUARTER)) %>%
  summarise(Sample = n(), Population = sum(PWT18)) 

N_total_2 %>% flextable() %>% theme_vanilla() %>%
  set_caption(caption = "Table 1: Sample size and population estimates", style = "Table Caption") %>%
  bold(bold = TRUE, part = "header") %>%
  set_header_labels(`as_label(QUARTER)` = "Quarter",
                    Population = "Population estimates")  %>%
  autofit() %>%
  align_nottext_col(align = "center", header = TRUE, footer = FALSE) %>%
  add_footer_lines("Source: UK Labour Force Survey (Person)") %>%
  color(part = "footer", color = "#666666")  

```


We focus on understanding how COVID-19 has impacted on the precarisation of work in the UK. Benach et al. (2016)^[Benach, J., Vives, A., Tarafa, G., Delclos, C., Muntaner, C.: What should we know about precarious employment and health in 2025? Framing the agenda for the next decade of research. Int. J. Epidemiol. 45(1), 232–238 (2016)] describe three approaches to defining precarious working common in existing research: (1) defining precariousness as relating to certain sectors of the labour market; (2) defining it as referring to any so called non-standard work type such as zero hours (i.e. any employment that is not through a permanent, full-time contract); and (3) defining it in relation to the attributes of a job, taking into account various aspects of the employment context. Although part-time contracts are actually quite ‘standard’ for women workers in the UK, we nevertheless follow the second approach and explore it in relation to gender, ethnicity and ocupational class. 

We follow the trend pre- and post-pandemic, considering March/May 2020 as the reference point. We trace the pre-pandemic trend up to the first 2019 quarter (Jan/March). We do this to identify if the COVID-19 pandemic is narrowing or reinforcing existing precarious work inequalities in the UK.

It is important to note that the LFS responses are weighted to official population projections that pre-date the COVID-19 pandemic. This particularly affects estimates for ethnicity. This implies that levels and differences in levels should be used with caution. Despite this, sample level estimates show a similar trend regarding gender differences for each group.


# For more information

* Luis D. Torres: luis.torres@nottingham.ac.uk; 
* Tracey Warren: tracey.warren@nottingham.ac.uk;
* Annegreet Veeken: annegreet.veeken@nottingham.ac.uk
*	[Project website](https://www.nottingham.ac.uk/business/research/carrying-the-work-burden-of-covid-19/index.aspx)


# Acknowledgements

*	The project ‘How is COVID-19 impacting women and men’s working lives in the UK? Expanding an existing award (on the work of working-class women in the UK) to explore all workers’ is part of the Data and Connectivity National Core Study, led by Health Data Research UK in partnership with the Office for National Statistics and funded by UK Research and Innovation (project ref HDRUK2020.137).

*	The research data are distributed by the UK Data Service (safeguarded data version). The original data creators, the depositors, the copyright holders, the funders and the UK Data Archive bear no responsibility for the analysis or interpretation of the data made in this paper.


# Citation

Torres L, Warren T,  Veeken A and the UK Women’s Budget Group (2021). How has the risk of precarious work evolved in the COVID-19 UK? Research Summary 2. Nottingham: Nottingham University Business School.



