---
title: "How has self-employment changed since the COVID-19 pandemic in the UK? Research Summary 3"
subtitle: 'Project: How is COVID-19 impacting women and men’s working lives in the UK?'
author: "Luis D. Torres, Tracey Warren and Annegreet Veeken, and the UK Women’s Budget Group"
date: August 2021
output: 
  word_document:
      reference_docx: "Report_template.docx"
      toc: FALSE
      toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
	warning = FALSE,
  collapse = TRUE)
```

```{r libraries}
#### packages needed
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(gridExtra)) install.packages("girdExtra", repos = "http://cran.us.r-project.org")
if(!require(corrr)) install.packages("corrr", repos = "http://cran.us.r-project.org")
if(!require(Hmisc)) install.packages("Hmisc", repos = "http://cran.us.r-project.org")
if(!require(sjlabelled)) install.packages("sjlabelled", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
if(!require(ggtext)) install.packages("ggtext", repos = "http://cran.us.r-project.org")
if(!require(ggrepel)) install.packages("ggrepel", repos = "http://cran.us.r-project.org")
if(!require(doBy)) install.packages("doBy", repos = "http://cran.us.r-project.org")
if(!require(sjmisc)) install.packages("sjmisc", repos = "http://cran.us.r-project.org")
if(!require(plotly)) install.packages("plotly", repos = "http://cran.us.r-project.org")
if(!require(flextable)) install.packages("flextable", repos = "http://cran.us.r-project.org")
if(!require(scales)) install.packages("scales", repos = "http://cran.us.r-project.org")
if(!require(ggalt)) install.packages("ggalt", repos = "http://cran.us.r-project.org")
if(!require(janitor)) install.packages("janitor", repos = "http://cran.us.r-project.org")
if(!require(bookdown)) install.packages("bookdown", repos = "http://cran.us.r-project.org")
if(!require(rmdformats)) install.packages("rmdformats", repos = "http://cran.us.r-project.org")
if(!require(htmlwidgets)) install.packages("htmlwidgets", repos = "http://cran.us.r-project.org")
if(!require(widgetframe)) install.packages("widgetframe", repos = "http://cran.us.r-project.org")
if(!require(shiny)) install.packages("shiny", repos = "http://cran.us.r-project.org")

library(ggalt) # contingency table
library(tidyverse) # several tools for data manipulation
library(haven) # read SPSS, Stata, and SAS data
library(gridExtra) # arrange multiple grid-based plots on a page
library(corrr) # for correlations correlate()
library(Hmisc) # many functions useful for data analysis
library(sjmisc) # useful to set Na's
library(sjlabelled) # get_labels function
library(ggthemes) # themes for ggplot2
library(ggtext) # fancy text
library(ggrepel) # help with the position of the labels on ggplot2 graph
library(doBy) # 'do' something to data stratified 'by' some variables
library(plotly) # interactive plots
library(flextable) # formatting tables 
library(scales) # add scale symbols and colour to axis
library(janitor) # add totals to tables
library(rmdformats)
library(htmlwidgets)
library(shiny)
library(widgetframe)

```

```{r data, echo=FALSE}

load("../../Data_clean/LFS_clean.rda") # this is the base file

d_self <- LFS_clean %>% # sud-dataset to facilitate analysis
  # set variable types
  mutate_at(c("QUARTER", "PWT18", "PIWT18"), ~as.numeric(.)) %>%  
  # Apply basic filters
  filter(AGE >= 18 & AGE <= 64) %>% 
  filter(ILODEFR == 1) # only people in employment 
```

# Background

Covid-19 has proven to be not solely a health crisis. Measures adopted by businesses and governments, such as flexible working and the job retention scheme, have shaped the impacts of the pandemic in ways not comparable to previous social and economic crises, while lockdowns and social distancing measures have had an unequal impact on those jobs frequently done by women and men. We set out to examine if Covid-19 is narrowing or reinforcing existing inequalities among women and men in their working lives. 

Our first and second research summaries focused on unemployment and precarious work, respectively. This research summary examines self-employment. The self-employed were predicted to be particularly vulnerable workers as the pandemic took hold, facing real threats to their livelihoods[^1].

[^1]:Enterprise Research Centre (2020) Covid-19 and self-employment in the UK. Retrieved from https://www.enterpriseresearch.ac.uk/wp-content/uploads/2020/04/ERC-Insight-Covid-19-and-self-employment-in-the-UK.pdf

Over 99% of all people in employment can be classified as employees or self-employed workers[^2] . Employees are those who are paid a wage by an employer for the work that they do. On the other hand, self-employed workers are those people who in their main employment work on their own account (with or without employees). The number of employees and self-employed people in the UK is measured by the Labour Force Survey (LFS). 

[^2]:Office of National Statistics (2020). A guide to labour market statistics. Retrieved from: https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/employmentandemployeetypes/methodologies/aguidetolabourmarketstatistics

This research summary aims at understanding how Covid-19 has impacted, if at all, on self-employment in the UK. To answer this question, we use the UK’s largest study on employment circumstances, the Labour Force Survey (LFS), drawing on analysis of the 2019/20/21 releases. We follow the trend pre- and post-pandemic, considering March/May 2020 as the reference point. We trace the pre-pandemic trend from the first 2019 quarter (Jan/March). We do this to identify if the Covid-19 pandemic is narrowing or reinforcing existing inequalities in self-employment trends for workers aged between 18-64.


# How did self-employment evolve after the pandemic hit?

```{r table self-employment total}

t_self_total <- d_self %>% # Totals
  # Apply basic filters
  group_by(QUARTER) %>% 
  summarise(self = sum(PWT18[INECAC05 == 2]))

```

The numbers of workers in self-employment fell after the start of 2020 (see Figure 1). Up to the end of 2019 self-employment had been increasing, a rise that started in the 1970s . However, the uncertainties that COVID-19 brought appear to have paused any short-term plans to become self-employed and made it difficult to continue in self-employment for those already working on their own account. 

Government support for the self-employed through the ‘Self-Employment Income Support Scheme’ helped to keep the numbers steady from May-July quarter 2020. However, the number of self-employed workers fell again from the October-December quarter, amid further uncertainties brought by the second lockdown in November 2020.


```{r plot self-employment total, fig.width= 8, fig.height= 5, fig.fullwidth=TRUE, fig.cap= "Figure 1: Self-employment dropped after the pandemic started"}

p_self_total <- t_self_total %>% 
  ggplot(aes(x = QUARTER, y = self)) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_vline(xintercept = 7, linetype="dotted") +
  geom_bar(stat="identity", width=0.5, fill = c("#367BB8")) + 
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,5000000, 500000), limits = c(0, 5000000), labels = comma) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
  caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 4800000, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 4800000, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 4800000, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "bottom", 
        legend.title = element_blank(),
        legend.text = element_text(size=11))

p_self_total # print 

```

The following sections present a detailed analysis of self-employment trends considering gender, ethnicity, and nation within the UK. To facilitate comparison, we calculate a self-employment rate. This is a percentage that summarises self-employment as a share of all workers (employees and self-employed). 


## Self-employment among women and men fell after the pandemic hit

```{r table self-employment rate sex}

t_self_rate <- d_self %>% # rate to in employment population
  # Apply basic filters
  filter(AGE >= 18 & AGE <= 64) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(empl = sum(PWT18[ILODEFR ==1]), self = sum(PWT18[INECAC05 == 2])) %>% 
  group_by(QUARTER, SEX) %>% 
  # rate compared to the 
  mutate(rate_self = self / empl)

```
Self-employment is highly gendered. There are far more self-employed males than there are females: women are less likely to set up in business, their operations tend to be smaller, and fewer self-employed women than men employ other workers.

Male workers were more likely to be self-employed than were women in all periods included in this analysis, with a gender gap of 8% in 2019. Men experienced the highest drops in post-pandemic self-employment levels: their rates fell from 18.2% at the end of 2019 to 15.2% at the beginning of 2021. Women saw a smaller decline from 10.6% to 9.1%.  


```{r plot self-employment rate sex, fig.width= 8, fig.height= 5, fig.fullwidth=TRUE, fig.cap= "Figure 2: Self-employed male workers numbers dropped the most"}

p_self_rate <- t_self_rate %>% 
  ggplot(aes(x = QUARTER, y = rate_self, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.208), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 17, y = 0.145, label = "Men", color= "#DF9216", size = 4) +
  annotate(geom = "text", x = 17, y = 0.085, label = "Women", color= "#791F83", size = 4) +
  annotate(geom = "text", x = 7.1, y = 0.03, label = "Start of the COVID-19\nfirst lockdown", hjust = 0, size = 3.1) +
  annotate(geom = "text", x = 2, y = 0.207, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.207, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.207, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "none") 

p_self_rate # plot

```

## Minority-ethnic self-employment rates fell the most

```{r table minority and sex}

d_self <- d_self %>% mutate(ETHUKEUL_2 = recode_factor(factor(ETHUKEUL), 
                                                             `1` = "White",
                                                             `2` = "Minority-ethnic", `3` = "Minority-ethnic", 
                                                             `4` = "Minority-ethnic", `5` = "Minority-ethnic",
                                                             `6` = "Minority-ethnic",`7` = "Minority-ethnic",
                                                             `8` = "Minority-ethnic",`9` = "Minority-ethnic")) %>% as_numeric(., keep.labels = TRUE)

t_self_rate_eth <- d_self %>% # rate to in employment population
  # Apply basic filters
  group_by(QUARTER, SEX, ETHUKEUL_2) %>% 
  summarise(empl = sum(PWT18[ILODEFR ==1]), self = sum(PWT18[INECAC05 == 2])) %>% 
  group_by(QUARTER, SEX, ETHUKEUL_2) %>% 
  # rate compared to the 
  mutate(rate = self / empl) %>%
  filter(!is.na(ETHUKEUL_2))

```
Ethnicity also plays a strong role in self-employment: men from minority-ethnic groups are more likely to be self-employed than are white men or minority-ethnic and white women workers. Restricted job opportunities and the prevalence of insecure work for minority-ethnic men are key contributing factors.

Men and women from minority-ethnic groups experienced the steepest post-pandemic falls in rates of self-employment. This is particularly the case for minority-ethnic men who saw a decline of 4% points (from 20% to 16%), while white men saw 3% points decrease (18% to 15%). White women experienced the smallest decrease of around 1% points, while ethnic-minority women saw a cut of 2% (from 11% to 9%). 


```{r plot minority and sex, fig.width= 12, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 3: Minority-ethnic self-employment rates fell the most."}

p_self_rate_eth <- t_self_rate_eth %>% 
  ggplot(aes(x = QUARTER, y = rate, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_grid(.~ as_label(ETHUKEUL_2)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.22, 0.02), limits = c(0, 0.225), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 17, y = 0.165, label = "Men", color= "#DF9216", size = 4) +
  annotate(geom = "text", x = 17, y = 0.105, label = "Women", color= "#791F83", size = 4) +
  annotate(geom = "text", x = 7.1, y = 0.03, label = "Start of the COVID-19\nfirst lockdown", hjust = 0, size = 3.1) +
  annotate(geom = "text", x = 2, y = 0.224, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.224, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.224, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "none", 
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size) 

p_self_rate_eth # plot
```

## Northern Ireland experienced the steepest fall in male self-employment

```{r table country and sex}

t_self_rate_sex_country <- d_self %>% # rate to in employment population
  # Apply basic filters
  group_by(QUARTER, COUNTRY2, SEX) %>% 
  summarise(empl = sum(PWT18[ILODEFR ==1]), self = sum(PWT18[INECAC05 == 2])) %>% 
  group_by(QUARTER, COUNTRY2, SEX) %>% 
  # rate compared to the 
  mutate(rate_self = self / empl)

```

All nations within the UK experienced the decline in male and female self-employment outlined above but, while England and Scotland followed the overall trend depicted in Figure 2, Wales and Northern Ireland showed a rather different pattern. 

In Wales, male self-employment rates had reached their highest levels at 19% in the second quarter of 2019, but then fell to their lowest rate in 2020 before the pandemic hit (14% in the first quarter of 2020). During the pandemic, rates remained between 15% and 16% and with some additional small recovery at the start of 2021. The picture is different for women whose self-employment remained above 8% until the start of the pandemic when it decreased to a low of 6% at the end of 2020. 

Male self-employment fell most steeply in Northern Ireland. Male rates here fell from 22% to 14% while female rates remained relatively stable.


```{r plot country and sex, fig.width= 12, fig.height= 7, fig.fullwidth=TRUE, fig.cap= "Figure 4: Northern Ireland experienced the steepest fall in male self-employment"}

p_self_rate_sex_country <- t_self_rate_sex_country %>% 
  ggplot(aes(x = QUARTER, y = rate_self, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(COUNTRY2)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.25, 0.02), limits = c(0, 0.25), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 9.5, y = 0.244, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.244, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size)  

p_self_rate_sex_country # plot

```

## Minority-ethnic self-employment fell dramatically in Scotland and Northern Ireland

```{r table country and ethnicity}

t_self_rate_eth_country <- d_self %>% # rate to in employment population
  # Apply basic filters
  group_by(QUARTER, COUNTRY2, ETHUKEUL_2) %>% 
  summarise(empl = sum(PWT18[ILODEFR ==1]), self = sum(PWT18[INECAC05 == 2])) %>% 
  group_by(QUARTER, COUNTRY2, ETHUKEUL_2) %>% 
  # rate compared to the 
  mutate(rate = self / empl) %>%
  filter(!is.na(ETHUKEUL_2))

```

Minority-ethnic self-employment rates showed significant variation among the four nations. While the self-employment rate was very similar for white and minority-ethnic workers in England; Wales, Scotland and Northern Ireland showed very different patterns. 

Wales experienced the highest levels of variability. Rates of self-employment among minority-ethnic workers reached their highest point at 20% in the third quarter of 2019 but fell to below 10% in the first quarter of 2020. Rates remained between 10% and 14% from the start of the pandemic until the end of 2020, returning to 2019 levels at the start of 2021. A similar trend can be seen among minority-ethnic workers in Northern Ireland. 

On the other hand, Scotland has experienced a persistent decrease in minority-ethnic self-employment rates. Levels fell from 17% in 2019 to a low of 4% at the end of 2020. As we entered 2021, rates went up to 6%. 


```{r plot country and ethnicity, fig.width= 12, fig.height= 7, fig.fullwidth=TRUE, fig.cap= "Figure 5: Minority-ethnic self-employment fell dramatically in Scotland and Northern Ireland"}

p_self_rate_eth_country <- t_self_rate_eth_country %>% 
  ggplot(aes(x = QUARTER, y = rate, colour = as_label(ETHUKEUL_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(COUNTRY2)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.20, 0.02), limits = c(0, 0.207), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.207, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.207, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.207, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "top", 
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size) 

p_self_rate_eth_country # plot

```


# How did the choice of first and second jobs evolve for men and women?

The LFS classifies a worker as employee or self-employed according to their main jobs. However, some workers also have a second job. This implies that someone who is an employee in their main job, could be self-employed in their second job and vice versa. 

In this section, we explore the relationship between main and second jobs with focus on self-employment. We first look at the second job options of those who are employees in the main job. We focus on self-employment as a second job. Then, we focus on the second job options of those who are self-employed in their main jobs. Here we explore if those in self-employment become employees or continue as self-employed in their second jobs This relationship can help us to understand the actual opportunities that self-employment offered to workers during the pandemic. For example, if employees are self-employed in a second job then this could be an indicator of self-employment as a positive option (e.g., complement income, explore new ideas, etc.), while a self-employed worker who is an employee in a second job could be a sign of struggles in their main job (e.g., uncertainties, insufficient income, etc). 


## Self-employment is more common as a second job for female employees

Is self-employment a second job option for those who are employees in their main jobs? The answer is yes, and particularly for women. Figure 6 shows that a higher number of female employees say that they are self-employed in their second job. This may be explained by the time flexibility that self-employment offers and their need to balance work with their own family responsibilities. The second semester of 2020 saw a further increase in self-employment as a second job for female employees.  

```{r table self-employment second job sex}

t_self_rate_sex_emain <- d_self %>% # main as an employee with a second job as self-employee
  filter(STATR == 1 & STAT2 == 2) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(total = sum(PWT18)) %>% 
  mutate(worker_type = 1)

t_self_rate_sex_smain <- d_self %>% # main as self-employee with a second job as  employee
  filter(STATR == 2 & STAT2 == 1) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(total = sum(PWT18)) %>% 
  mutate(worker_type = 2)

t_self_rate_sex_smains <- d_self %>% # main as self-employee with a second job as self-employee
  filter(STATR == 2 & STAT2 == 2) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(total = sum(PWT18)) %>% 
  mutate(worker_type = 3)

t_self_rate_sex_main_second <- rbind(t_self_rate_sex_emain, t_self_rate_sex_smain, t_self_rate_sex_smains)

t_self_rate_sex_main_second <- t_self_rate_sex_main_second %>% set_labels(worker_type, 
                                                                          labels = c("Main job: Employee | Second job: Self-employed" = 1,
                                                                                     "Main job: Self-employed | Second job: Employee" = 2,
                                                                                     "Main job: Self-employed | Second job: Self-employed" = 3))

t_self_rate_sex_main_second <- t_self_rate_sex_main_second %>% set_labels(SEX, 
                                                                          labels = c("Men" = 1,
                                                                                     "Women" = 2))

```


```{r plot self-employment second job sex, fig.width= 8, fig.height= 5, fig.fullwidth=TRUE, fig.cap= "Figure 6: Self-employment is more common as a second job for women"}

p_self_rate_sex_main_second  <- t_self_rate_sex_main_second  %>% filter(worker_type == 1) %>%
  ggplot(aes(x = QUARTER, y = total, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(worker_type)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,220000, 20000), limits = c(0, 227000), labels = comma) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Number of workers",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 227000, label = "2019", size = 3.1) +  
  annotate(geom = "text", x = 9.5, y = 227000, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 227000, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 9),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size)  

p_self_rate_sex_main_second # plot


```

## Being an employee in a second job is more common among self-employed women than men

What about the second jobs options of those who are self-employed in their main jobs? Figure 7 shows the second job options of those who are self-employed in their main jobs. The figure shows two second job options for those in self-employment: switch to employee and continue as self-employed. The left side of the figure shows that self-employed women tend to switch back to being employees in their second jobs. The pandemic reduced this trend for most of 2020, most likely because of less employment opportunities during the pandemic. However, as we entered 2021 numbers rose once again.

By contrary, more self-employed men than women tend to choose self-employment also as their second job (right side on Figure 7). This trend saw a reduction since the pandemic hit until the end of 2020. The beginning of 2021 showed signs of a new increase for male and female self-employed workers. 

```{r plot self-employment second job sex 2, fig.width= 12, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 7: Being an employee in a second job is more common among self-employed women than men"}

p_self_rate_sex_main_second_2  <- t_self_rate_sex_main_second  %>% filter(worker_type != 1) %>%
  ggplot(aes(x = QUARTER, y = total, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~as_label(worker_type)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,100000, 10000), limits = c(0, 105000), labels = comma) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Number of workers",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 105000, label = "2019", size = 3.1) +  
  annotate(geom = "text", x = 9.5, y = 105000, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 105000, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 9),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size)  

p_self_rate_sex_main_second_2 # plot

```




# Did the capacity of the self-employed to hire employees change? 

```{r table selfemployed with employees overall}

t_self_main_overall <- d_self %>% 
  filter(STATR == 2) %>%
  group_by(QUARTER) %>% 
  summarise(total = sum(PWT18)) # total self employment in main job 

t_self_main_employees_overall <- d_self %>% 
  filter(STATR == 2 & SOLOR == 2) %>% 
  group_by(QUARTER) %>% 
  summarise(with_employees = sum(PWT18)) # total self-employed with employees in main job

t_self_main_employees_rate_overall <- left_join(t_self_main_overall, t_self_main_employees_overall, by = c("QUARTER"))
t_self_main_employees_rate_overall <- t_self_main_employees_rate_overall  %>% mutate(rate = with_employees / total, category = 1)

t_self_second_overall <- d_self %>% 
  filter(STAT2 == 2) %>%
  group_by(QUARTER) %>% 
  summarise(total = sum(PWT18)) # total self employment in second job

t_self_second_employees_overall <- d_self %>% 
  filter(STAT2 == 2 & SOLO2 == 2) %>% 
  group_by(QUARTER) %>% 
  summarise(with_employees = sum(PWT18)) # total self-employed with employees in second job

t_self_second_employees_rate_overall <- left_join(t_self_second_overall, t_self_second_employees_overall, by = c("QUARTER"))
t_self_second_employees_rate_overall <- t_self_second_employees_rate_overall  %>% mutate(rate = with_employees / total, category = 2)

t_self_employees_rate_overall <- rbind(t_self_main_employees_rate_overall, t_self_second_employees_rate_overall)

t_self_employees_rate_overall <- t_self_employees_rate_overall %>% set_labels(category, 
                                                                labels = c("Self-employment as main job" = 1,
                                                                           "Self-employment as second job" = 2))
```


The deep uncertainties being experienced by the self-employed had ramifications for the wider labour force. We ask next whether self-employed workers were hiring staff. Figure 8 shows that they were having trouble to do to. Since the pandemic hit the number of self-employed workers having employees progressively declined and continued doing so as we entered 2021. The decline was particularly sharp for those self-employed workers in their second job (6% points fall). Some recovery was being experienced by these workers as we entered 2021.


```{r plot selfemployed with employees overall, fig.width= 12, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 8: Self-employed with employees in their main and second jobs"}


p_self_employees_overall <- t_self_employees_rate_overall %>% 
  ggplot(aes(x = QUARTER, y = rate, col = "#360EA9")) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(category)) + 
  
  # scale and colour
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.208), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Percentage with employees",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.207, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.207, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.207, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 10),
        legend.position = "none", 
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size) 

p_self_employees_overall # plot

```


## Self-employed women are less likely to have employees

```{r table selfemployed with employees rate}

t_self_main <- d_self %>% 
  filter(STATR == 2) %>%
  group_by(QUARTER, SEX) %>% 
  summarise(total = sum(PWT18)) # total self employment in main job 

t_self_main_employees <- d_self %>% 
  filter(STATR == 2 & SOLOR == 2) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(with_employees = sum(PWT18)) # total self-employed with employees in main job

t_self_main_employees_rate <- left_join(t_self_main, t_self_main_employees, by = c("QUARTER", "SEX"))
t_self_main_employees_rate <- t_self_main_employees_rate  %>% mutate(rate = with_employees / total, category = 1)

t_self_second <- d_self %>% 
  filter(STAT2 == 2) %>%
  group_by(QUARTER, SEX) %>% 
  summarise(total = sum(PWT18)) # total self employment in second job

t_self_second_employees <- d_self %>% 
  filter(STAT2 == 2 & SOLO2 == 2) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(with_employees = sum(PWT18)) # total self-employed with employees in second job

t_self_second_employees_rate <- left_join(t_self_second, t_self_second_employees, by = c("QUARTER", "SEX"))
t_self_second_employees_rate <- t_self_second_employees_rate  %>% mutate(rate = with_employees / total, category = 2)

t_self_employees_rate <- rbind(t_self_main_employees_rate, t_self_second_employees_rate)

t_self_employees_rate <- t_self_employees_rate %>% set_labels(category, 
                                                                labels = c("Self-employment as main job" = 1,
                                                                           "Self-employment as second job" = 2))
```

Figure 9 shows that, overall, self-employed women are less likely than are their male counterparts to have employees working for them. There was a clear gender gap in whether they were an employer for both types of self-employed (self-employed in their main and second jobs).

Since the pandemic hit the number of self-employed women having employees progressively declined and continued doing so as we entered 2021. This trend applied both to those women who are in self-employment in their main jobs as well as those who are self-employed in their second jobs. Self-employed men in their main and second jobs both experienced a decline in hiring employees since the pandemic hit, but it was particularly abrupt for those self-employed in their second job (8% points fall). Some recovery was being experienced by these workers as we entered 2021. 


```{r plot selfemployed with employees rate, fig.width= 12, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 9: Self-employed women are less likely to have employees"}


p_self_employees_rate <- t_self_employees_rate %>% 
  ggplot(aes(x = QUARTER, y = rate, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(category)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.2, 0.02), limits = c(0, 0.208), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Percentage with employees",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.207, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.207, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.207, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 10),
        legend.position = "top", 
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size) 

p_self_employees_rate # plot

```

## Falling numbers of self-employed minority-ethnic workers are employers

```{r table selfemployed with employees rate by ethnicity}

t_self_main_eth <- d_self %>% 
  filter(STATR == 2) %>%
  group_by(QUARTER, ETHUKEUL_2) %>% 
  summarise(total = sum(PWT18)) # total self employment in main job 

t_self_main_employees_eth <- d_self %>% 
  filter(STATR == 2 & SOLOR == 2) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  summarise(with_employees = sum(PWT18)) # total self-employed with employees in main job

t_self_main_employees_rate_eth <- left_join(t_self_main_eth, t_self_main_employees_eth, by = c("QUARTER", "ETHUKEUL_2"))
t_self_main_employees_rate_eth <- t_self_main_employees_rate_eth  %>% mutate(rate = with_employees / total, category = 1)

t_self_second_eth <- d_self %>% 
  filter(STAT2 == 2) %>%
  group_by(QUARTER, ETHUKEUL_2) %>% 
  summarise(total = sum(PWT18)) # total self employment in second job

t_self_second_employees_eth <- d_self %>% 
  filter(STAT2 == 2 & SOLO2 == 2) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  summarise(with_employees = sum(PWT18)) # total self-employed with employees in second job

t_self_second_employees_rate_eth <- left_join(t_self_second_eth, t_self_second_employees_eth, by = c("QUARTER", "ETHUKEUL_2"))
t_self_second_employees_rate_eth <- t_self_second_employees_rate_eth %>% mutate(rate = with_employees / total, category = 2)

t_self_employees_rate_eth <- rbind(t_self_main_employees_rate_eth, t_self_second_employees_rate_eth) %>% filter(!is.na(ETHUKEUL_2))

t_self_employees_rate_eth <- t_self_employees_rate_eth %>% set_labels(category, 
                                                              labels = c("Self-employment as main job" = 1,
                                                                         "Self-employment as second job" = 2))
t_self_employees_rate_eth <- t_self_employees_rate_eth %>% set_labels(ETHUKEUL_2, 
                                                                      labels = c("White" = 1,
                                                                                 "Minority-ethnic" = 2))


```

Self-employed minority-ethnic workers are more likely to have employees working for them than are the white self-employed (Figure 10). 
Since the pandemic hit the number of self-employed minority-ethnic workers having employees progressively declined through 2020. The start of 2021 did show signs of recovery, but only for those self-employed in their main jobs. The situation was different for those minority-ethnic workers who were self-employed as their second jobs: the percentage having employees dropped dramatically (from over 16% to below 2%). 


```{r plot selfemployed with employees rate by ethnicity, fig.width= 12, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 10: Self-employed minority-ethnic workers are less likely to employ workers after the pandemic hit"}

p_self_employees_rate_eth <- t_self_employees_rate_eth %>% 
  ggplot(aes(x = QUARTER, y = rate, colour = as_label(ETHUKEUL_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_wrap(.~ as_label(category)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +
  scale_y_continuous(breaks = seq(0,0.25, 0.02), limits = c(0, 0.25), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Percentage with employees",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.23, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.23, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.23, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 9),
        axis.text.y = element_text(size = 10),
        legend.position = "top", 
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size) 

p_self_employees_rate_eth # plot


```

# How have the hours worked by the self-employed changed?

```{r table hous}

hours <- d_self %>% select(QUARTER, INECAC05, SEX, ETHUKEUL_2, PAIDHRU, BUSHR, TTUSHR, EVEROT, POTHR, UOTHR) %>%
  filter(INECAC05 %in% c(1, 2))

```

The self-employed report real diversity in how many hours they spend on their businesses a week, with some working only a few hours and others spending far longer at work. How many hours do self-employed people spend in work per week and did this change with the pandemic? 

Figure 11 shows the range and distribution of the total usual hours (usual hours and any additional worked per week) for self-employed workers. The main impact of the pandemic has been in the distribution of typical and maximum hours. By looking at the size of the box in the figure we can see that typical hours in pre-pandemic 2020 were concentrated around 27 to 45 hours, maximum hours of between 70 to 73 hours per week, and a median of 40 hours. Since the pandemic hit typical paid hours were concentrated between 25 to 45 hours per week, maximum hours up to 75, while the median remained at 40 hours. This has been relatively stable since the pandemic hit. 


```{r plot hours overall, fig.width= 8, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 11: Total usual hours worked per week for all quarters"}

p_hours_overall <- hours %>%  filter(!is.na(TTUSHR) & INECAC05 == 2) %>%
  ggplot(aes(x = as_factor(QUARTER), y = TTUSHR, fill = as_factor(QUARTER))) + 
  geom_boxplot(alpha=0.3, outlier.shape=TRUE, show.legend = TRUE) +
  geom_vline(xintercept = 4.2, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15.5, color = "#c9c9c9") + # gridline between 2020/2021
  geom_vline(xintercept = 7.2, linetype="dotted") + 
  
  # scale and colour
  coord_cartesian(ylim = c(0, 100)) +
  scale_y_continuous(breaks = seq(0,100, 5)) + 
  scale_fill_manual(values = c("#360EA9", "#360EA9", "#360EA9", "#360EA9", "#360EA9" ,"#360EA9" ,"#360EA9", "#360EA9", "#360EA9", "#360EA9", "#360EA9", "#360EA9","#360EA9", "#360EA9", "#360EA9", "#360EA9", "#360EA9" , "#360EA9")) +
  scale_x_discrete(limits=c("1","2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18"),
                   labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                            "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                            "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                            "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Usual hours per week - main job",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 103, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 103, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 103, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 11),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "none", 
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 11)) + # change the face_wrap title
  guides(col = guide_legend(override.aes = list(size = 5)))

p_hours_overall # plot

```



## Self-employed women work fewer paid hours than do men and their hours vary the most

Self-employed women tend to spend shorter weeks at work than do men. Many of the women work in their business on a part-time basis, fewer than 30 hours a week, while the male self-employed can report working very long weeks. Figure 12 compares the range and distribution of the total usual hours for male and female self-employed workers. By looking at the size of the box in the figure we can see that usual hours for female self-employed workers vary greatly compared to men. 

Since the pandemic hit typical total usual hours for men have been concentrated around 35 and 45 hours per week, while women’s range between 16 to 40 hours per week. Median paid hours remained stable for both: at 40 per week for men and 30 for women. 


```{r plot hours, fig.width= 8, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 12: Self-employed women work fewer hours per week"}

p_hours <- hours %>%  filter(!is.na(TTUSHR) & INECAC05 == 2) %>%
  ggplot(aes(x = as_factor(QUARTER), y = TTUSHR, fill = as_label(SEX))) + 
  geom_boxplot(aes(ymin=..lower.., ymax=..upper..), alpha=0.3, outlier.shape=NA, show.legend = TRUE) +
  # stat_summary(aes(col = as_label(SEX)), fun = mean, geom = "point", shape=19, size = 3, show.legend = TRUE, alpha=1) + 
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_vline(xintercept = 7, linetype="dotted") + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +  
  scale_fill_manual(values = c("#DF9216", "#791F83")) + 
  coord_cartesian(ylim = c(0, 60)) +
  scale_y_continuous(breaks = seq(0,60, 5)) + 
  scale_x_discrete(limits=c("1","2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18"),
                   labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                            "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                            "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                            "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Usual hours per week - main job",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 58, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 58, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 58, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 11),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top", 
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 11)) + # change the face_wrap title
  guides(col = guide_legend(override.aes = list(size = 5)))

p_hours# plot

```

```{r table self-employment part-time}

t_self <- d_self %>% 
  filter(INECAC05 == 2) %>%
  group_by(QUARTER, SEX) %>% 
  summarise(total = sum(PWT18))

t_self_ft <- d_self %>% # rate to in employment population
  filter(INECAC05 == 2 & FTPTWK == 1) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(subtotal= sum(PWT18)) %>%
  mutate(jobtype = 1)

t_self_pt <- d_self %>% # rate to in employment population
  filter(INECAC05 == 2 & FTPTWK == 2) %>% 
  group_by(QUARTER, SEX) %>% 
  summarise(subtotal= sum(PWT18)) %>%
  mutate(jobtype = 2)

t_self_ft_rate <- left_join(t_self, t_self_ft, by = c("QUARTER", "SEX"))
t_self_ft_rate <- t_self_ft_rate %>% mutate(rate = subtotal / total)

t_self_pt_rate <- left_join(t_self, t_self_pt, by = c("QUARTER", "SEX"))
t_self_pt_rate <- t_self_pt_rate %>% mutate(rate = subtotal / total)

t_self_ftpt_rate <- rbind(t_self_ft_rate, t_self_pt_rate) 

t_self_ftpt_rate <- t_self_ftpt_rate %>% set_labels(jobtype, 
                                                           labels = c("Full-time" = 1,
                                                                      "Part-time" = 2))

```




```{r plot self-employment part-time, fig.width= 8, fig.height= 5, fig.fullwidth=TRUE, fig.cap= "Figure 13: Half of self-employed women work part-time"}

p_self_ftpt_rate <- t_self_ftpt_rate %>% filter(jobtype == 2) %>% 
  ggplot(aes(x = QUARTER, y = rate, colour = as_label(SEX))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_grid(.~ as_label(jobtype)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +  
  scale_y_continuous(breaks = seq(0,0.6, 0.05), limits = c(0, 0.6), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.58, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.58, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.58, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "top", 
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size) 

p_self_ftpt_rate # plot


```


## The ethnic gap in paid hours among the self-employed is narrowing

Average hours worked a week fluctuated from around 38 to 40 hours for both groups of the self-employed, minority-ethnic and white, and in all quarters examined. Typical hours for white self-employed workers ranged from around 25 to 45 hours per week, while minority-ethnic workers worked between 25 to 40 hours per week. 

The gap in the number of total hours worked by white and minority-ethnic self-employed workers narrowed after the end of 2020. 


```{r plot hours ethnicity, fig.width= 8, fig.height= 6, fig.fullwidth=TRUE, fig.cap= "Figure 14: The usual hours gap between white and minority-ethnic self-employed workers narrowed"}


p_hours_eth <- hours %>%  filter(!is.na(ETHUKEUL_2) & !is.na(TTUSHR) & INECAC05 == 2) %>% 
  ggplot(aes(x = as_factor(QUARTER), y = TTUSHR, fill = as_label(ETHUKEUL_2))) + 
  geom_boxplot(aes(ymin=..lower.., ymax=..upper..), alpha=0.3, outlier.shape=NA, show.legend = TRUE) +
  #stat_summary(aes(col = as_label(ETHUKEUL_2)), fun = mean, geom = "point", shape=19, size = 3, show.legend = TRUE, alpha=1) + 
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_vline(xintercept = 7, linetype="dotted") + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +  
  scale_fill_manual(values = c("#DF9216", "#791F83")) + 
  coord_cartesian(ylim = c(0, 60)) +
  scale_y_continuous(breaks = seq(0,60, 5)) + 
  scale_x_discrete(limits=c("1","2","3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18"),
                   labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                            "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                            "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                            "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "Usual hours per week - main job",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 58, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 58, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 58, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 11),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        legend.position = "top", 
        legend.title = element_blank(),
        legend.text = element_text(size = 11),
        strip.text = element_text(face = "bold", hjust = 0, size = 11)) + # change the face_wrap title
  guides(col = guide_legend(override.aes = list(size = 5)))

p_hours_eth # plot


```


```{r table part-time ethnicity}

t_self_eth <- d_self %>% 
  filter(INECAC05 == 2) %>%
  group_by(QUARTER, ETHUKEUL_2) %>% 
  summarise(total = sum(PWT18))

t_self_ft_eth <- d_self %>% # rate to in employment population
  filter(INECAC05 == 2 & FTPTWK == 1) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  summarise(subtotal= sum(PWT18)) %>%
  mutate(jobtype = 1)

t_self_pt_eth <- d_self %>% # rate to in employment population
  filter(INECAC05 == 2 & FTPTWK == 2) %>% 
  group_by(QUARTER, ETHUKEUL_2) %>% 
  summarise(subtotal= sum(PWT18)) %>%
  mutate(jobtype = 2)

t_self_ft_rate_eth <- left_join(t_self_eth, t_self_ft_eth, by = c("QUARTER", "ETHUKEUL_2"))
t_self_ft_rate_eth <- t_self_ft_rate_eth %>% mutate(rate = subtotal / total)

t_self_pt_rate_eth <- left_join(t_self_eth, t_self_pt_eth, by = c("QUARTER", "ETHUKEUL_2"))
t_self_pt_rate_eth <- t_self_pt_rate_eth %>% mutate(rate = subtotal / total)

t_self_ftpt_rate_eth <- rbind(t_self_ft_rate_eth, t_self_pt_rate_eth) 

t_self_ftpt_rate_eth <- t_self_ftpt_rate_eth %>% set_labels(jobtype, 
                                                    labels = c("Full-time" = 1,
                                                               "Part-time" = 2)) %>% 
  filter(!is.na(ETHUKEUL_2)) %>% 
  tibble(.)

t_self_ftpt_rate_eth <- t_self_ftpt_rate_eth %>% set_labels(ETHUKEUL_2, 
                                                            labels = c("White" = 1,
                                                                       "Minority-ethnic" = 2)) %>% tibble(.)

```


```{r plot part-time ethnicity, fig.width= 8, fig.height= 5, fig.fullwidth=TRUE, fig.cap= "Figure 15: The ethnicity gap in part-time is narrowing"}

p_self_ftpt_rate_eth <- t_self_ftpt_rate_eth %>% filter(jobtype == 2) %>% 
  ggplot(aes(x = as.numeric(QUARTER), y = as.numeric(rate), colour = as_label(ETHUKEUL_2))) + # line graph
  geom_vline(xintercept = 4, color = "#c9c9c9") + # gridline between 2019/2020
  geom_vline(xintercept = 15, color = "#c9c9c9") + # gridline between 2020/2021
  geom_line(size = 1, show.legend = FALSE) +
  geom_point(shape = 19) +
  geom_vline(xintercept = 7, linetype="dotted") +
  facet_grid(.~ as_label(jobtype)) + 
  
  # scale and colour
  scale_colour_manual(values = c("#DF9216", "#791F83")) +  
  scale_y_continuous(breaks = seq(0,0.4, 0.05), limits = c(0, 0.4), labels =
                       scales::percent_format(accuracy = 1L)) +
  scale_x_continuous(breaks=c(1:18),
                     labels=c("Jan-\nMar","Apr-\nJun","Jul-\nSep","Oct-\nDec",
                              "Jan-\nMar", "Feb-\nApr", "Mar-\nMay", "Apr-\nJun", "May-\nJul",
                              "Jun-\nAug", "Jul-\nSep","Aug-\nOct", "Sep-\nNov", "Oct-\nDec",
                              "Nov-\nJan", "Dec-\nFeb", "Jan-\nMar", "Feb-\nApr")) +
  # annotation
  labs(x = "", 
       y = "",
       caption = c("Source: UK Labour Force Survey (Person)")) +
  annotate(geom = "text", x = 2, y = 0.38, label = "2019", size = 3.1) +
  annotate(geom = "text", x = 9.5, y = 0.38, label = "2020", size = 3.1) +
  annotate(geom = "text", x = 17, y = 0.38, label = "2021", size = 3.1) +
  
  # theme
  theme_economist_white(gray_bg = FALSE) +
  theme(plot.margin = unit(c(1, 5, 1, 1), "lines"),
        plot.title = element_text(margin = ggplot2::margin(10, 0, 10, 0)),
        plot.subtitle = element_text(size = 12, hjust = 0),
        axis.title.x = element_text(margin = ggplot2::margin(t = 5),
                                    vjust = 0, size = 12),
        axis.title.y = element_text(margin = ggplot2::margin(r = 3),
                                    vjust = 2, size = 12),
        axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 10),
        legend.position = "top", 
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        strip.text = element_text(face = "bold", hjust = 0, size = 12)) + # change the face_wrap title
  guides(color = guide_legend(nrow = 1, byrow = TRUE, override.aes = list(size = 4))) # change the legend box size) 

p_self_ftpt_rate_eth # plot

```




# Conclusions and Recommendations

## Conclusions 

We examined in this report if the Covid-19 pandemic impacted the self-employed in the UK. We looked at women and men, the type of self-employment, whether the self-employed were employers or not, and their hours worked per week.

The self-employed were seen to be particularly vulnerable workers as the pandemic took hold, facing real threats to their livelihoods. Indeed the numbers of self-employed fell from over 4.4 million at the start of 2019 to 3.8 million by April 2021 (Table 1). These workers are over concentrated in sectors like domestic services, personal care, and accommodation and food that were shut down fully or severely limited with the pandemic restrictions, and government support schemes for the self-employed were less well-known and excluded those newly self-employed.

Drawing on analysis of 2019/20/21 data, we showed that the numbers of self-employed workers fell across the UK, with some variations in England, Scotland, Wales and Northern Ireland. The impact of the pandemic on self-employment is not random but gendered, ethnically, and geographically driven. Self-employed people were less able to hire staff, with ramifications for unemployment levels among the labour force, and their own working hours varied more than they did pre the pandemic, with potential disruption to their incomes and financial security. 

We showed the importance of looking at gender in an analysis of self-employment but also the necessity of including ethnicity and nation when we consider who is self-employed and why as the pandemic rolled out. Who is self-employed and the risks that they faced during the pandemic are not random but highly dependent on gender, ethnicity and how they intersect.


## Recommendations

In order to improve the working lives of everyone, particularly in the post-pandemic context, policies must take into account the fact that gender and ethnicity impact the likelihood of a person to be in self-employment. 

To be effective, policies to support the self-employed should therefore tackle the root causes of structural inequalities, rather than simply address their manifestations.


# The Project and its Data

<p align="justify">

Our project *'How is Covid-19 impacting women and men's working lives in the UK?'* is funded by Health Data Research UK, as part of the rapid funding call to use and enrich the data within the Data & Connectivity National Core Study (NCS) capability. This report draws on the analysis of the 2019/20/21 releases of the UK Labour Force Survey (LFS). Respondents are interviewed for five successive waves at three-months intervals and approximately 20% of the sample is replaced every quarter. Four quarters releases are supported in a typical year: Jan-Mar \| Apr-Jun \| Jul-Sep \| Oct-Dec. From 2020, additional non-calendar quarter data have been released in response to the context of the coronavirus pandemic. Besides the four calendar quarters per year, 2020 included another eight releases of data. The last two quarters in 2020 also included data for January and February 2021.

</p>

<p align="justify">

Our analyses consider only those in employment (current jobs) between 18-64 years old. Table 1 summarises the sample size and population estimates for all the data waves considered in this analysis.

</p>

```{r sample, echo=FALSE}

## CREATING A SUB-DATASET FOR ANALYSIS

d_self <- LFS_clean %>% 
  # set variable types
  #mutate_at(c("QUARTER", "PWT18", "PIWT18"), ~as.numeric(.)) %>%  
  # Apply basic filters
  filter(AGE >= 18 & AGE <= 64) %>% 
  filter(ILODEFR == 1) # only people in employment 

## Sample and population   
  
N_total_3 <- d_self %>%
  group_by(as_label(QUARTER)) %>%
  summarise(Sample = n(), In_employment = sum(PWT18), Self_employed = sum(PWT18[INECAC05 == 2])) 

N_total_3 %>% flextable() %>% theme_vanilla() %>%
  set_caption(caption = "Table 1: Sample size and population estimates", style = "Table Caption") %>%
  bold(bold = TRUE, part = "header") %>%
  set_header_labels(`as_label(QUARTER)` = "Quarter",
                    In_employment = "In employment",
                    Self_employed = " Self-employed")  %>%
  autofit() %>%
  align_nottext_col(align = "center", header = TRUE, footer = FALSE) %>%
  add_footer_lines("Note: Workers between 18-64 years old") %>%
  add_footer_lines("Source: UK Labour Force Survey (Person)") %>%
  color(part = "footer", color = "#666666")  

```

<p align="justify">

It is important to note that the LFS responses are weighted to official population projections that pre-date the Covid-19 pandemic. This particularly affects estimates for ethnicity. This implies that levels and differences in levels should be used with caution. Despite this, sample level estimates show a similar trend regarding gender differences for each group.

</p>

# For more information

## Contact details

-   Luis D. Torres: [luis.torres\@nottingham.ac.uk](mailto:luis.torres@nottingham.ac.uk){.email};
-   Tracey Warren: [tracey.warren\@nottingham.ac.uk](mailto:tracey.warren@nottingham.ac.uk){.email};
-   Annegreet Veeken: [annegreet.veeken\@nottingham.ac.uk](mailto:annegreet.veeken@nottingham.ac.uk){.email}
-   The UK Women's Budget Group: [sara.reis\@wbg.org.uk](mailto:sara.reis@wbg.org.uk){.email}
-   [Project website](https://www.nottingham.ac.uk/business/research/carrying-the-work-burden-of-covid-19/index.aspx)


## Acknowledgements

-   <p align="justify">The project 'How is COVID-19 impacting women and men's working lives in the UK? Expanding an existing award (on the work of working-class women in the UK) to explore all workers' is part of the Data and Connectivity National Core Study, led by Health Data Research UK in partnership with the Office for National Statistics and funded by UK Research and Innovation (project ref HDRUK2020.137).</p>

-   <p align="justify"> The research data are distributed by the UK Data Service (safeguarded data version). The original data creators, the depositors, the copyright holders, the funders and the UK Data Archive bear no responsibility for the analysis or interpretation of the data made in this paper.</p>


## Citation

<p align="justify"> Torres L, Warren T, Veeken A and the UK Women's Budget Group (2021). *How has self-employment changed since the COVID-19 pandemic in the UK?* Research Summary 3. Nottingham: Nottingham University Business School. </p>
